---
title: "An Extra Sum of Stress: Does Economic Distress Predict Democratic Support Beyond General Prosperity and Demographic Composition?"
author: "Khanh Do, Maral Bat-Erdene"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggrepel)
library(readr)
library(tidyverse)
library(corrplot)
require(leaps)
library(patchwork)
require(mosaic)
require(corrplot)
require(leaps)
library(ggplot2)
library(reshape2)
```

# Read in Data

```{r message = FALSE, warning = FALSE, echo = FALSE}
url <- "https://raw.githubusercontent.com/khanhdo05/sta-310-final/main/data/clean/FINAL_DATA.csv"
df <- read_csv(url, show_col_types = FALSE)
```

Read csv from this [link](https://raw.githubusercontent.com/khanhdo05/sta-310-final/main/data/clean/FINAL_DATA.csv) into a dataframe `df`.

Code book available [here](https://docs.google.com/spreadsheets/d/1A-s4d9eTGT6rGAFhPsAnOutXermwbm6LxZr9JXc_NPE/edit?gid=0#gid=0).

Each row is unique by column `FIPS`.

# 1. Data Transformation

## Response Variable: PresDemRatio

```{r echo=FALSE}
hist(df$PresDemRatio,
     main = "Distribution of PresDemRatio",
     col = "steelblue")
```

The data looks right-skewed, but we shouldn't use log-transformation because this variable is a proportion bounded by 0 and 1. Log-transformation is too aggressive for this data and can turn a right-skewed distribution to left-skewed. A square-root transformation is milder and more appropriate. We can see this more clearly with the histograms below.

```{r echo=FALSE}
hist(log(df$PresDemRatio),
     main = "Distribution of Log-Transformed PresDemRatio",
     col = "steelblue")
hist(sqrt(df$PresDemRatio),
     main = "Distribution of Sqrt-Transformed PresDemRatio",
     col = "steelblue")
```

The distribution of the square-root transformed one is not perfect but is definitely more bell-shape looking than that with log. Thus, we determine that our response variable needs a square-root transformation.

## Exploratory Variable

```{r}
# ---------------- PREP DATA ----------------
df_out <- df %>%
  select(-FIPS, -CountyName) %>%
  drop_na()
response   <- "PresDemRatio"
predictors <- setdiff(names(df_out), response)

# ---------------- OUTLIER DETECTION METHOD: COOK'S DISTANCE ----------------
# ---------------- LOOP THROUGH VARIABLES ----------------
par(mfrow = c(1,1))
for (var in predictors) {
  
  temp <- df_out %>%
    select(all_of(c(response, var))) %>%
    rename(x = all_of(var), y = all_of(response)) %>%
    drop_na()
  
  n <- nrow(temp)
  
  # Cook's Distance outlier detection
  model   <- lm(y ~ x, data = temp)
  cooks_d <- cooks.distance(model)
  cutoff  <- 4 / n
  
  temp <- temp %>%
    mutate(
      cooks_d    = cooks_d,
      is_outlier = cooks_d > cutoff
    )
  
  # ---------------- PRINT VARIABLE HEADER ----------------
  cat("\n---------------------------------------------\n")
  cat("Variable:", var, "\n")
  cat("Number of influential points (Cook's D > 4/n):", sum(temp$is_outlier), "\n")
  
  # ---------------- PLOT ----------------
  p <- ggplot(temp, aes(x = x, y = y)) +
    geom_point(alpha = 0.5, color = "gray40") +
    
    # Highlight influential points
    geom_point(
      data = temp %>% filter(is_outlier),
      color = "red",
      size = 2
    ) +
    
    # Label the most influential ones
    geom_text_repel(
      data = temp %>% filter(is_outlier) %>% slice_max(cooks_d, n = 5),
      aes(label = round(x, 2)),
      size = 3
    ) +
    
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    
    labs(
      title    = paste("Influential Point Detection:", var, "vs PresDemRatio"),
      subtitle = paste0("Cook's D threshold: 4/n = ", round(cutoff, 4),
                        "  |  Influential points: ", sum(temp$is_outlier)),
      x = var,
      y = "PresDemRatio"
    ) +
    theme_minimal()
  
  print(p)
}
```

Cook's Distance (Cook, 1977) is used here as the primary outlier detection method because it measures the influence of each observation on the fitted regression coefficients which is directly relevant to our goal of building a regression model. Unlike distributional methods such as IQR or Z-score outlier detection, which only flagged observations based on a single variable's distribution, Cook's D identifies observations whose removal would substantially change the estimated relationship between a predictor and the response variable (Belsley, Kuh & Welsch, 1980). The standard threshold of 4/n (Cook & Weisberg, 1982) is applied, where n is the number of complete observations for each predictor-response pair.

```{r}
library(tidyverse)
library(corrplot)
library(leaps)
library(car)

# ---------------- PREP DATA ----------------
df <- df %>%
  select(-FIPS, -CountyName) %>%
  drop_na()

cat("Sample size:", nrow(df), "\n")
```


```{r}
# ---------------- CORRELATION PLOT ----------------
# Remove categorical/binary variables before computing correlation
df_corr <- df %>%
  select(-IsMetro, -PresDemRatio) %>%
  select(where(is.numeric))

corrplot(cor(df_corr),
         title = "Correlation Matrix (Final Data)", 
         mar = c(0,0,1,0), tl.cex = 0.5)
```

The correlation matrix shows that all income and tax dollar-amount variables (NumTaxReturns through ChildTaxCredit) form a near-perfectly correlated block, the age variables (MedianAge, Pct65Plus, OldAgeDependency, ChildDependency) cluster together, and the education variables (PctNoHs through PctBachelors) seems a little intercorrelated. Since our goal is purely predictive rather than inferential, we will not address multicollinearity and instead allow the best subsets selection procedure to identify the combination of variables that best predicts PresDemRatio.


```{r}
# ---------------- PAIRS PLOTS (GROUPED) ----------------

# Group 1: Income / Economic
pairs(df %>% select(PresDemRatio, TotalAgi, SalariesWages, BusinessNetIncome,
                    CapitalGains, UnemploymentComp, SocialSecurityBenefits),
      main = "Income Variables vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("steelblue", alpha.f = 0.4))

# Group 2: Education & Demographics
pairs(df %>% select(PresDemRatio, PctNoHs, PctHsOnly, PctSomeCollege,
                    PctBachelors, PctWhite, PctBlack, PctHispanic),
      main = "Education & Race vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("darkgreen", alpha.f = 0.4))

# Group 3: Population & Age
pairs(df %>% select(PresDemRatio, CensusPop2020, MedianAge, Pct18_34,
                    Pct65Plus, AgeDependency),
      main = "Population & Age vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("purple", alpha.f = 0.4))

# Group 4: Taxes & Wealth
pairs(df %>% select(PresDemRatio, TotalTaxLiability, MortgageInterest,
                    CharitableContributions, EicAmount, NumEicReturns),
      main = "Tax & Wealth vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("firebrick", alpha.f = 0.4))
```


```{r}
# ---------------- MISSING DATA ----------------
missing_summary <- df %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing") %>%
  mutate(
    Total = nrow(df),
    Pct_Missing = round(100 * Missing / Total, 1)
  ) %>%
  arrange(desc(Missing))

print(missing_summary)

# ---------------- BEST SUBSETS ----------------
df$sq_PresDemRatio <- sqrt(df$PresDemRatio)

regfit <- regsubsets(sq_PresDemRatio ~ ., data = df, nvmax = 25)

plot(regfit, scale = "adjr2",
     main = "Best Subsets (Voting Model)",
     cex.axis = 0.6)

get_best <- function(model) {
  s <- summary(model)
  data.frame(
    NumVars = 1:length(s$adjr2),
    AdjR2 = s$adjr2,
    s$which
  ) %>% arrange(desc(AdjR2))
}

best_models <- get_best(regfit)
head(best_models, 10)

# ---------------- FIT MODEL ----------------
lm_best <- lm(sq_PresDemRatio ~ 
                TotalAgi +
                CapitalGains +
                UnemploymentRate2020 +
                PctBachelors +
                PctWhite +
                MedianAge +
                CensusPop2020 +
                TotalTaxLiability +
                EicAmount,
              data = df)

summary(lm_best)

# ---------------- RESIDUAL DIAGNOSTICS ----------------
par(mfrow = c(1,1))
plot(lm_best)

# ---------------- RESIDUALS VS EACH VARIABLE ----------------
vars <- names(df)
vars <- vars[vars != "sq_PresDemRatio"]

par(mfrow = c(1,1))

for (v in vars) {
  plot(df[[v]], lm_best$residuals,
       xlab = v,
       ylab = "Residuals",
       main = paste("Residuals vs", v),
       pch = 16, cex = 0.5)
  abline(h = 0, col = "red", lty = 2)
}
```

