---
title: "An Extra Sum of Stress: Does Economic Distress Predict Democratic Support Beyond General Prosperity and Demographic Composition?"
author: "Khanh Do, Maral Bat-Erdene"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggrepel)
library(readr)
library(tidyverse)
library(corrplot)
require(leaps)
library(patchwork)
require(mosaic)
require(corrplot)
require(leaps)
library(ggplot2)
library(reshape2)
library(car)
```

# Read in Data

```{r message = FALSE, warning = FALSE, echo = FALSE}
url <- "https://raw.githubusercontent.com/khanhdo05/sta-310-final/main/data/clean/FINAL_DATA.csv"
df <- read_csv(url, show_col_types = FALSE)
```

Read csv from this [link](https://raw.githubusercontent.com/khanhdo05/sta-310-final/main/data/clean/FINAL_DATA.csv) into a dataframe `df`.

Code book available [here](https://docs.google.com/spreadsheets/d/1A-s4d9eTGT6rGAFhPsAnOutXermwbm6LxZr9JXc_NPE/edit?gid=0#gid=0).

Each row is unique by column `FIPS`.

```{r echo=FALSE}
cat("Original data's nubmer of rows:", nrow(df), "\n")
cat("Original data's nubmer of columns:", ncol(df), "\n")
```


# 1. Data Transformation

## 1.1 Response Variable: PresDemRatio

```{r echo=FALSE}
hist(df$PresDemRatio,
     main = "Distribution of PresDemRatio",
     col = "steelblue")
```

The data looks right-skewed, but we shouldn't use log-transformation because this variable is a proportion bounded by 0 and 1. Log-transformation is too aggressive for this data and can turn a right-skewed distribution to left-skewed. A square-root transformation is milder and more appropriate. We can see this more clearly with the histograms below.

```{r echo=FALSE}
hist(log(df$PresDemRatio),
     main = "Distribution of Log-Transformed PresDemRatio",
     col = "steelblue")
hist(sqrt(df$PresDemRatio),
     main = "Distribution of Sqrt-Transformed PresDemRatio",
     col = "steelblue")
```

The distribution of the square-root transformed one is not perfect but is definitely more bell-shape looking than that with log. Thus, we determine that our response variable needs a square-root transformation.

```{r echo=FALSE}
# Apply sqrt transformation to response
df$sq_PresDemRatio <- sqrt(df$PresDemRatio)
```


## 1.2 Exploratory Variable

### 1.2.1 Check for Missing Data

```{r echo=FALSE}
# ---------------- MISSING DATA ----------------
missing_summary <- df %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing") %>%
  mutate(
    Total = nrow(df),
    Pct_Missing = round(100 * Missing / Total, 1)
  ) %>%
  arrange(desc(Missing))

print(missing_summary)

# remove rows with NAs
df1 <- df %>%
  drop_na()
cat("Number of rows before removing rows with NAs:", nrow(df), "\n")
cat("Number of rows after removing rows with NAs:", nrow(df1), "\n")
```

None of our variables have more than 0.5% of missing data so we do not have to exclude any variables (remove no columns).

### 1.2.2 Outliers

```{r echo=FALSE}
# Prep data
df_out <- df %>%
  select(-FIPS, -CountyName) %>%
  drop_na()

response   <- "PresDemRatio"
predictors <- setdiff(names(df_out), response)

# Create a list to store plots
plot_list <- list()

# Outlier detection loop
for (var in predictors) {
  
  temp <- df_out %>%
    select(all_of(c(response, var))) %>%
    rename(x = all_of(var), y = all_of(response))
  
  n <- nrow(temp)
  model   <- lm(y ~ x, data = temp)
  cooks_d <- cooks.distance(model)
  cutoff  <- 4 / n
  
  temp <- temp %>%
    mutate(cooks_d = cooks_d, is_outlier = cooks_d > cutoff)
  
  # Create the ggplot and save it to the list
  p <- ggplot(temp, aes(x = x, y = y)) +
    geom_point(aes(color = is_outlier), alpha = 0.4) +
    scale_color_manual(values = c("gray60", "red")) +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    labs(
      title = paste(var),
      subtitle = paste("Outliers:", sum(temp$is_outlier)),
      x = NULL, y = NULL
    ) +
    theme_minimal(base_size = 8) +
    theme(legend.position = "none")
  
  plot_list[[var]] <- p
}

# Display plots
wrap_plots(plot_list[1:9], ncol = 3, nrow = 3) +
  plot_annotation(title = "Outlier Detection 1")
wrap_plots(plot_list[10:18], ncol = 3, nrow = 3) +
  plot_annotation(title = "Outlier Detection 2")
wrap_plots(plot_list[19:27], ncol = 3, nrow = 3) +
  plot_annotation(title = "Outlier Detection 3")
wrap_plots(plot_list[28:36], ncol = 3, nrow = 3) +
  plot_annotation(title = "Outlier Detection 4")
wrap_plots(plot_list[37:39], ncol = 1, nrow = 3) +
  plot_annotation(title = "Outlier Detection 5")
```

Cook's Distance (Cook, 1977) is used here as the primary outlier detection method because it measures the influence of each observation on the fitted regression coefficients which is directly relevant to our goal of building a regression model. Unlike distributional methods such as IQR or Z-score outlier detection, which only flagged observations based on a single variable's distribution, Cook's D identifies observations whose removal would substantially change the estimated relationship between a predictor and the response variable (Belsley, Kuh & Welsch, 1980). The standard threshold of 4/n (Cook & Weisberg, 1982) is applied, where n is the number of complete observations for each predictor-response pair. Although the outliers are identified, we don't remove them without good reasons.

### 1.2.3 Correlation Plots

```{r echo=FALSE}
# Remove categorical/binary variables before computing correlation
df_corr <- df_out %>%
  select(-IsMetro, -PresDemRatio) %>%
  select(where(is.numeric))

corrplot(cor(df_corr),
         title   = "Correlation Matrix",
         mar     = c(0, 0, 1, 0),
         tl.cex  = 0.5)
```

The correlation matrix shows that all income and tax dollar-amount variables (NumTaxReturns through ChildTaxCredit) form a near-perfectly correlated block, the age variables (MedianAge, Pct65Plus, OldAgeDependency, ChildDependency) cluster together, and the education variables (PctNoHs through PctBachelors) are moderately intercorrelated. Since our initial analysis goal is strong predictive rather than inferential, we will not address multicollinearity and instead allow the best subsets selection procedure to identify the combination of variables that best predicts PresDemRatio. Multicollinearity will be addressed formally in part **1.2.4**. 


```{r echo=FALSE}
# ---------------- PAIRS PLOTS (GROUPED) ----------------
# Group 1: Income / Economic
pairs(df_out %>% select(PresDemRatio, TotalAgi, SalariesWages, PropSalaries, AvgWage, BusinessNetIncome,
                    CapitalGains, UnemploymentComp, SocialSecurityBenefits),
      main = "Income Variables vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("steelblue", alpha.f = 0.4))

# Group 2: Education & Demographics
pairs(df_out %>% select(PresDemRatio, PctNoHs, PctHsOnly, PctSomeCollege,
                    PctBachelors, PctWhite, PctBlack, PctHispanic),
      main = "Education & Race vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("darkgreen", alpha.f = 0.4))

# Group 3: Population & Age
pairs(df_out %>% select(PresDemRatio, CensusPop2020, MedianAge, Pct18_34,
                    Pct65Plus, AgeDependency),
      main = "Population & Age vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("purple", alpha.f = 0.4))

# Group 4: Taxes & Wealth
pairs(df_out %>% select(PresDemRatio, TotalTaxLiability, MortgageInterest,
                    CharitableContributions, EicAmount, NumEicReturns),
      main = "Tax & Wealth vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("firebrick", alpha.f = 0.4))
```

**Income Variables vs Voting:**

All income variables (TotalAgi, SalariesWages, CapitalGains, UnemploymentComp, SocialSecurityBenefits) show a heavily right-skewed, L-shaped relationship with PresDemRatio, with nearly all data points compressed near zero, suggesting these variables need log-transformation before modeling. 

PropSalaries and AvgWage look slightly skewed and would benefit from square-root transformation.

**Education & Race vs Voting:**

PctBachelors has a visible positive trend with PresDemRatio, while PctNoHs and PctWhite show negative trends. PctBlack and PctHispanic show positive but noisier relationships. These variables seem to be the most promising predictors in their current form with no transformation needed.

**Population & Age vs Voting:**

CensusPop2020 is extremely right-skewed and needs log-transformation. MedianAge, Pct18_34, and Pct65Plus show weak, noisy relationships with PresDemRatio individually, though Pct65Plus and MedianAge looks correlated with each other.

**Tax & Wealth vs Voting:** 

L-shaped compression also show in TotalTaxLiability, MortgageInterest, CharitableContributions, EicAmount, and NumEicReturns, and thus all need log-transformation. Also, EicAmount and NumEicReturns looks correlated with each other.


```{r echo=FALSE}
# ---------------- APPLY TRANSFORMATIONS ----------------
df2 <- df %>%
  mutate(
    # Log-transform the financial variables
    log_TotalAgi                = log(TotalAgi + 1),
    log_SalariesWages           = log(SalariesWages + 1),
    log_BusinessNetIncome       = log(BusinessNetIncome + 1),
    log_CapitalGains            = log(CapitalGains + 1),
    log_UnemploymentComp        = log(UnemploymentComp + 1),
    log_SocialSecurityBenefits  = log(SocialSecurityBenefits + 1),
    log_TotalTaxLiability       = log(TotalTaxLiability + 1),
    log_MortgageInterest        = log(MortgageInterest + 1),
    log_CharitableContributions = log(CharitableContributions + 1),
    log_EicAmount               = log(EicAmount + 1),
    log_CensusPop2020           = log(CensusPop2020 + 1),
    
    # sqrt-transform lightly skewed data
    sqrt_PropSalaries           = sqrt(PropSalaries),
    sqrt_AvgWage                = sqrt(AvgWage) 
  ) %>%
  # Drop original untransformed versions only, keep everything else
  select(
    -TotalAgi, -SalariesWages, -BusinessNetIncome, -CapitalGains,
    -UnemploymentComp, -SocialSecurityBenefits, -TotalTaxLiability,
    -MortgageInterest, -CharitableContributions, -EicAmount, -CensusPop2020,
    -PresDemRatio, -PropSalaries, -AvgWage
  )

# ---------------- VERIFY TRANSFORMATIONS ----------------
par(mfrow = c(3, 4))
log_vars <- c("log_TotalAgi", "log_SalariesWages", "log_BusinessNetIncome",
              "log_CapitalGains", "log_UnemploymentComp", "log_SocialSecurityBenefits",
              "log_TotalTaxLiability", "log_MortgageInterest",
              "log_CharitableContributions", "log_EicAmount", "log_CensusPop2020",
              "sqrt_PropSalaries", "sqrt_AvgWage")

for (v in log_vars) {
  hist(df2[[v]],
       main = v, xlab = "", col = "steelblue", cex.main = 0.8)
}
par(mfrow = c(1, 1))

cat("Number of columns now:", ncol(df2), "\n")
cat("Number of rows now:", nrow(df2), "\n")
```
### 1.2.4 Multicollinearity (VIF) and mean-centering

```{r}
# TODO: IsMetro is not working.

# Build full candidate model to compute VIFs
df_vif <- df_best %>%
  select(
    # Response
    sq_PresDemRatio,
    
    # Income (log-transformed)
    log_TotalAgi, log_SalariesWages, log_BusinessNetIncome,
    log_CapitalGains, log_UnemploymentComp, log_SocialSecurityBenefits,
    log_TotalTaxLiability, log_MortgageInterest,
    log_CharitableContributions, log_EicAmount, log_CensusPop2020,
    
    # Other income
    PropSalaries, AvgWage,
    
    # Education (PctNoHs dropped as reference)
    PctHsOnly, PctSomeCollege, PctBachelors,
    
    # Race (PctWhite dropped as reference)
    PctBlack, PctAsian, PctHispanic,
    
    # Demographics
    PctMale, PctUrban, IsMetro,
    
    # Age
    MedianAge, Pct18_34, Pct65Plus,
    AgeDependency, OldAgeDependency, ChildDependency,
    
    # Other (NO duplicates)
    UnemploymentRate2020, NetMigration2020,
    BirthRate2020, DeathRate2020,
    AvgWage, PropSalaries,
    NumTaxReturns, PartnershipIncome,
    NumEicReturns, ChildTaxCredit
    
    # REMOVED: PctNoHs, PctWhite (said dropped)
  ) %>%
  drop_na()

# IsMetro (aliased coefficients in the model issue)
df_vif <- df_vif %>% select(-IsMetro)

#Create full model
lm_full_vif <- lm(sq_PresDemRatio ~ ., data = df_vif)
# Run VIF
vif_vals <- vif(lm_full_vif)
vif_df   <- data.frame(Variable = names(vif_vals), VIF = round(vif_vals, 2)) %>%
  arrange(desc(VIF))
print(vif_df)

# Flag high-VIF variables (more than 10 is standard threshold)
cat("\nHigh VIF variables:\n")
print(vif_df %>% filter(VIF > 10))
```

Our VIF analysis shows extreme multicollinearity in the full model, with several predictors having extraordinarily large VIF values that indicate strong relationships among variables. 

In particular, income-related variables such as *log_TotalAgi* (VIF = 10,384.37) and *log_SalariesWages* (VIF = 9,678.03) are almost perfectly collinear, thus only one of these variables should be kept. 

Similarly, scale variables such as *log_CensusPop2020* (VIF = 559.23) and *log_TotalTaxLiability* (VIF = 554.50) are strongly tied to *NumTaxReturns* (VIF = 39.94), suggesting one representative measure (ex: population) should be kept while the others are removed. 

Also, *log_EicAmount* (VIF = 77.27), *NumEicReturns* (VIF = 67.40), and *ChildTaxCredit* (VIF = 58.91) are seems redundant, so only one should be included. 

Age-related predictors also display structural multicollinearity, with *OldAgeDependency* (VIF = 141.65), *Pct65Plus* (VIF = 113.30), *ChildDependency* (VIF = 23.31), and *MedianAge* (VIF = 22.73), thus a single summary measure such as *MedianAge* or *Pct65Plus* should be retained while dependency ratios are excluded. 

Lastly, *PropSalaries* (VIF = 45.78) and *AvgWage* (VIF = 14.92) are both related to labor composition and in broader income components such as *log_BusinessNetIncome* (VIF = 19.80) and *log_CapitalGains* (VIF = 26.15). 

```{r}
# ---------------- MEAN-CENTERING ----------------
# Mean-centering is essential before introducing interaction terms in Step 5
# to reduce multicollinearity between main effects and their interactions
# and to make intercepts interpretable

df_model <- df_vif %>%
  mutate(across(
    -c(sq_PresDemRatio),  # don't center response
    ~ . - mean(., na.rm = TRUE),
    .names = "{.col}"              # overwrite in place (centered)
  ))

cat("Mean-centering confirmed:\n")
print(round(colMeans(df_model %>% select(-sq_PresDemRatio)), 4))

# Sanity check
cat("Response variable range:\n")
print(summary(df_model$sq_PresDemRatio))
cat("Variables in model data:", ncol(df_model) - 1, "\n")
cat("Sample size:", nrow(df_model), "\n")
```

## Best Subsets - Two Groups

```{r echo=FALSE}
# ---------------- PREP FOR BEST SUBSETS ----------------
# TODO: Adding correct IsMetro for further analysis in group 1

# Group 1: Prosperity & Demographic Controls
df_group1 <- df_model %>%
  select(
    sq_PresDemRatio,
    log_TotalAgi, log_SalariesWages, log_CapitalGains,
    log_SocialSecurityBenefits, log_MortgageInterest,
    log_CharitableContributions, log_TotalTaxLiability,
    log_CensusPop2020, PropSalaries, AvgWage,
    PctBachelors, PctHsOnly, PctSomeCollege,
    PctBlack, PctAsian, PctHispanic,
    PctMale, PctUrban,
    MedianAge, Pct18_34,
    NetMigration2020, BirthRate2020, DeathRate2020
  )

# Group 2: Economic Distress Variables
df_group2 <- df_model %>%
  select(
    sq_PresDemRatio,
    UnemploymentRate2020,
    log_UnemploymentComp,
    log_EicAmount,
    NumEicReturns,
    ChildTaxCredit,
    OldAgeDependency,
    ChildDependency,
    log_BusinessNetIncome
  )
```


```{r echo=FALSE}
# ---------------- BEST SUBSETS: GROUP 1 ----------------
fit_g1 <- regsubsets(
  sq_PresDemRatio ~ .,
  data   = df_group1,
  nvmax  = 20,
  method = "seqrep"
)

s_g1 <- summary(fit_g1)

best_models_g1 <- data.frame(
  NumVars = 1:length(s_g1$adjr2),
  AdjR2   = round(s_g1$adjr2, 4),
  s_g1$which
) %>% arrange(desc(AdjR2))

cat("=== GROUP 1: Prosperity & Demographics ===\n")
head(best_models_g1, 10)

best_n_g1 <- which.max(s_g1$adjr2)
cat("Best model size (Adj R2):", best_n_g1, "variables\n")
cat("Best Adj R2:", round(max(s_g1$adjr2), 4), "\n")

# Plot
plot(s_g1$adjr2, type = "b",
     xlab = "Number of Variables",
     ylab = "Adjusted R2",
     main = "Group 1 — Adjusted R2")
points(best_n_g1, max(s_g1$adjr2), col = "red", pch = 16, cex = 1.5)

plot(fit_g1, scale = "adjr2",
     main = "Group 1: Variable Inclusion",
     cex.axis = 0.55)

# Selected variables
vars_g1 <- names(coef(fit_g1, best_n_g1))[-1]
cat("Selected from Group 1:\n")
print(vars_g1)
```

```{r}
# ---------------- BEST SUBSETS: GROUP 2 ----------------
fit_g2 <- regsubsets(
  sq_PresDemRatio ~ .,
  data   = df_group2,
  nvmax  = 10,
  method = "seqrep"
)

s_g2 <- summary(fit_g2)

best_models_g2 <- data.frame(
  NumVars = 1:length(s_g2$adjr2),
  AdjR2   = round(s_g2$adjr2, 4),
  s_g2$which
) %>% arrange(desc(AdjR2))

cat("\n=== GROUP 2: Economic Distress ===\n")
head(best_models_g2, 10)

best_n_g2 <- which.max(s_g2$adjr2)
cat("Best model size (Adj R2):", best_n_g2, "variables\n")
cat("Best Adj R2:", round(max(s_g2$adjr2), 4), "\n")

# Plot
plot(s_g2$adjr2, type = "b",
     xlab = "Number of Variables",
     ylab = "Adjusted R2",
     main = "Group 2 — Adjusted R2")
points(best_n_g2, max(s_g2$adjr2), col = "red", pch = 16, cex = 1.5)

plot(fit_g2, scale = "adjr2",
     main = "Group 2: Variable Inclusion",
     cex.axis = 0.55)

# Selected variables
vars_g2 <- names(coef(fit_g2, best_n_g2))[-1]
cat("Selected from Group 2:\n")
print(vars_g2)
```

```{r}
# ---------------- REDUCED MODEL (GROUP 1 ONLY) ----------------
reduced_formula <- as.formula(
  paste("sq_PresDemRatio ~", paste(vars_g1, collapse = " + "))
)

lm_reduced <- lm(reduced_formula, data = df_model)

# ---------------- RESIDUAL DIAGNOSTICS ----------------

cat("\n=== REDUCED MODEL SUMMARY ===\n")
summary(lm_reduced)

par(mfrow = c(1, 1))
plot(lm_reduced)
par(mfrow = c(1, 1))

# # ---------------- RESIDUALS VS EACH PREDICTOR ----------------
# model_df <- model.frame(lm_reduced)
# res <- residuals(lm_reduced)
# for (v in vars_g1) {
#   plot(model_df[[v]], res,
#        xlab = v,
#        ylab = "Residuals",
#        main = paste("Residuals vs", v),
#        pch  = 16, cex = 0.5,
#        col  = adjustcolor("steelblue", alpha.f = 0.5))
#   
#   abline(h = 0, col = "red", lty = 2)
# }
```

```{r echo=FALSE}
# ---------------- FIT BEST MODEL ----------------
# TODO
# ---------------- RESIDUAL DIAGNOSTICS ----------------
# TODO
# # ---------------- RESIDUALS VS EACH PREDICTOR ----------------
# for (v in vars_best) {
#   plot(df_model[[v]], lm_best$residuals,
#        xlab = v,
#        ylab = "Residuals",
#        main = paste("Residuals vs", v),
#        pch  = 16, cex = 0.5,
#        col  = adjustcolor("steelblue", alpha.f = 0.5))
#   abline(h = 0, col = "red", lty = 2)
# }
```

