---
title: "An Extra Sum of Stress: Does Economic Distress Predict Democratic Support Beyond General Prosperity and Demographic Composition?"
author: "Khanh Do, Maral Bat-Erdene"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggrepel)
library(readr)
library(tidyverse)
library(corrplot)
require(leaps)
library(patchwork)
require(mosaic)
require(corrplot)
require(leaps)
library(ggplot2)
library(reshape2)
```

# Read in Data

```{r message = FALSE, warning = FALSE, echo = FALSE}
url <- "https://raw.githubusercontent.com/khanhdo05/sta-310-final/main/data/clean/FINAL_DATA.csv"
df <- read_csv(url, show_col_types = FALSE)
```

Read csv from this [link](https://raw.githubusercontent.com/khanhdo05/sta-310-final/main/data/clean/FINAL_DATA.csv) into a dataframe `df`.

Code book available [here](https://docs.google.com/spreadsheets/d/1A-s4d9eTGT6rGAFhPsAnOutXermwbm6LxZr9JXc_NPE/edit?gid=0#gid=0).

Each row is unique by column `FIPS`.

# 1. Data Transformation

## Response Variable: PresDemRatio

```{r echo=FALSE}
hist(df$PresDemRatio,
     main = "Distribution of PresDemRatio",
     col = "steelblue")
```

The data looks right-skewed, but we shouldn't use log-transformation because this variable is a proportion bounded by 0 and 1. Log-transformation is too aggressive for this data and can turn a right-skewed distribution to left-skewed. A square-root transformation is milder and more appropriate. We can see this more clearly with the histograms below.

```{r echo=FALSE}
hist(log(df$PresDemRatio),
     main = "Distribution of Log-Transformed PresDemRatio",
     col = "steelblue")
hist(sqrt(df$PresDemRatio),
     main = "Distribution of Sqrt-Transformed PresDemRatio",
     col = "steelblue")
```

The distribution of the square-root transformed one is not perfect but is definitely more bell-shape looking than that with log. Thus, we determine that our response variable needs a square-root transformation.

## Exploratory Variable

```{r}
# ---------------- PREP DATA ----------------
df_out <- df %>%
  select(-FIPS, -CountyName) %>%
  drop_na()
response   <- "PresDemRatio"
predictors <- setdiff(names(df_out), response)

# ---------------- OUTLIER DETECTION METHOD: COOK'S DISTANCE ----------------
# ---------------- LOOP THROUGH VARIABLES ----------------
par(mfrow = c(1,1))
for (var in predictors) {
  
  temp <- df_out %>%
    select(all_of(c(response, var))) %>%
    rename(x = all_of(var), y = all_of(response)) %>%
    drop_na()
  
  n <- nrow(temp)
  
  # Cook's Distance outlier detection
  model   <- lm(y ~ x, data = temp)
  cooks_d <- cooks.distance(model)
  cutoff  <- 4 / n
  
  temp <- temp %>%
    mutate(
      cooks_d    = cooks_d,
      is_outlier = cooks_d > cutoff
    )
  
  # ---------------- PRINT VARIABLE HEADER ----------------
  cat("\n---------------------------------------------\n")
  cat("Variable:", var, "\n")
  cat("Number of influential points (Cook's D > 4/n):", sum(temp$is_outlier), "\n")
  
  # ---------------- PLOT ----------------
  p <- ggplot(temp, aes(x = x, y = y)) +
    geom_point(alpha = 0.5, color = "gray40") +
    
    # Highlight influential points
    geom_point(
      data = temp %>% filter(is_outlier),
      color = "red",
      size = 2
    ) +
    
    # Label the most influential ones
    geom_text_repel(
      data = temp %>% filter(is_outlier) %>% slice_max(cooks_d, n = 5),
      aes(label = round(x, 2)),
      size = 3
    ) +
    
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    
    labs(
      title    = paste("Influential Point Detection:", var, "vs PresDemRatio"),
      subtitle = paste0("Cook's D threshold: 4/n = ", round(cutoff, 4),
                        "  |  Influential points: ", sum(temp$is_outlier)),
      x = var,
      y = "PresDemRatio"
    ) +
    theme_minimal()
  
  print(p)
}
```

Cook's Distance (Cook, 1977) is used here as the primary outlier detection method because it measures the influence of each observation on the fitted regression coefficients which is directly relevant to our goal of building a regression model. Unlike distributional methods such as IQR or Z-score outlier detection, which only flagged observations based on a single variable's distribution, Cook's D identifies observations whose removal would substantially change the estimated relationship between a predictor and the response variable (Belsley, Kuh & Welsch, 1980). The standard threshold of 4/n (Cook & Weisberg, 1982) is applied, where n is the number of complete observations for each predictor-response pair.

```{r}
# ---------------- CORRELATION PLOT ----------------
# Remove categorical/binary variables before computing correlation
df_corr <- df_out %>%
  select(-IsMetro, -PresDemRatio) %>%
  select(where(is.numeric))

corrplot(cor(df_corr),
         title = "Correlation Matrix (Final Data)", 
         mar = c(0,0,1,0), tl.cex = 0.5)
```

The correlation matrix shows that all income and tax dollar-amount variables (NumTaxReturns through ChildTaxCredit) form a near-perfectly correlated block, the age variables (MedianAge, Pct65Plus, OldAgeDependency, ChildDependency) cluster together, and the education variables (PctNoHs through PctBachelors) seems a little intercorrelated. Since our goal is purely predictive rather than inferential, we will not address multicollinearity and instead allow the best subsets selection procedure to identify the combination of variables that best predicts PresDemRatio.


```{r}
# ---------------- PAIRS PLOTS (GROUPED) ----------------

# Group 1: Income / Economic
pairs(df_out %>% select(PresDemRatio, TotalAgi, SalariesWages, BusinessNetIncome,
                    CapitalGains, UnemploymentComp, SocialSecurityBenefits),
      main = "Income Variables vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("steelblue", alpha.f = 0.4))

# Group 2: Education & Demographics
pairs(df_out %>% select(PresDemRatio, PctNoHs, PctHsOnly, PctSomeCollege,
                    PctBachelors, PctWhite, PctBlack, PctHispanic),
      main = "Education & Race vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("darkgreen", alpha.f = 0.4))

# Group 3: Population & Age
pairs(df_out %>% select(PresDemRatio, CensusPop2020, MedianAge, Pct18_34,
                    Pct65Plus, AgeDependency),
      main = "Population & Age vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("purple", alpha.f = 0.4))

# Group 4: Taxes & Wealth
pairs(df_out %>% select(PresDemRatio, TotalTaxLiability, MortgageInterest,
                    CharitableContributions, EicAmount, NumEicReturns),
      main = "Tax & Wealth vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("firebrick", alpha.f = 0.4))
```

Income Variables vs Voting:

All income variables (TotalAgi, SalariesWages, BusinessNetIncome, CapitalGains, UnemploymentComp, SocialSecurityBenefits) show a heavily right-skewed, L-shaped relationship with PresDemRatio, with nearly all data points compressed near zero, suggesting these variables need log-transformation before modeling.

Education & Race vs Voting: 

PctBachelors has a visible positive trend with PresDemRatio, while PctNoHs and PctWhite show negative trends. PctBlack and PctHispanic show positive but noisier relationships. These variables seem to be the most promising predictors in their current form with no transformation needed.

Population & Age vs Voting:

CensusPop2020 is extremely right-skewed and needs log-transformation. MedianAge, Pct18_34, and Pct65Plus show weak, noisy relationships with PresDemRatio individually, though Pct65Plus and MedianAge looks correlated with each other.

Tax & Wealth vs Voting: 

L-shaped compression also show in TotalTaxLiability, MortgageInterest, CharitableContributions, EicAmount, and NumEicReturns, and thus all need log-transformation. Also, EicAmount and NumEicReturns looks correlated with each other.

```{r}
# ---------------- MISSING DATA ----------------
missing_summary <- df %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing") %>%
  mutate(
    Total = nrow(df),
    Pct_Missing = round(100 * Missing / Total, 1)
  ) %>%
  arrange(desc(Missing))

print(missing_summary)
```


```{r}
# ---------------- PREP FOR BEST SUBSETS ----------------
df$sq_PresDemRatio <- sqrt(df$PresDemRatio)

# ---------------- TRANSFORMATIONS ----------------
df_best <- df %>%
  mutate(
    # Log-transform the financial variables
    log_TotalAgi                = log(TotalAgi + 1),
    log_SalariesWages           = log(SalariesWages + 1),
    log_BusinessNetIncome       = log(BusinessNetIncome + 1),
    log_CapitalGains            = log(CapitalGains + 1),
    log_UnemploymentComp        = log(UnemploymentComp + 1),
    log_SocialSecurityBenefits  = log(SocialSecurityBenefits + 1),
    log_TotalTaxLiability       = log(TotalTaxLiability + 1),
    log_MortgageInterest        = log(MortgageInterest + 1),
    log_CharitableContributions = log(CharitableContributions + 1),
    log_EicAmount               = log(EicAmount + 1),
    log_CensusPop2020           = log(CensusPop2020 + 1)
  ) %>%
  # Drop original untransformed versions only, keep everything else
  select(
    -TotalAgi, -SalariesWages, -BusinessNetIncome, -CapitalGains,
    -UnemploymentComp, -SocialSecurityBenefits, -TotalTaxLiability,
    -MortgageInterest, -CharitableContributions, -EicAmount, -CensusPop2020,
    -PresDemRatio
  )

# ---------------- VERIFY TRANSFORMATIONS ----------------
par(mfrow = c(3, 4))
log_vars <- c("log_TotalAgi", "log_SalariesWages", "log_BusinessNetIncome",
              "log_CapitalGains", "log_UnemploymentComp", "log_SocialSecurityBenefits",
              "log_TotalTaxLiability", "log_MortgageInterest",
              "log_CharitableContributions", "log_EicAmount", "log_CensusPop2020")

for (v in log_vars) {
  hist(df_best[[v]],
       main = v, xlab = "", col = "steelblue", cex.main = 0.8)
}
par(mfrow = c(1, 1))

cat("Remaining variables:", ncol(df_best), "\n")
cat("Sample size:", nrow(df_best), "\n")
print(names(df_best))

# Drop one category from each compositional group to remove linear dependencies
# Education: drop PctNoHs (reference), Race: drop PctWhite (reference)
df_model <- df_best %>%
  select(
    sq_PresDemRatio,
    
    # Income (log-transformed)
    log_TotalAgi, log_SalariesWages, log_BusinessNetIncome,
    log_CapitalGains, log_UnemploymentComp, log_SocialSecurityBenefits,
    log_TotalTaxLiability, log_MortgageInterest,
    log_CharitableContributions, log_EicAmount, log_CensusPop2020,
    
    # Education (PctNoHs dropped as reference)
    PctHsOnly, PctSomeCollege, PctBachelors,
    
    # Race (PctWhite dropped as reference)
    PctBlack, PctAsian, PctHispanic,
    
    # Demographics
    PctMale, PctUrban, IsMetro,
    
    # Age
    MedianAge, Pct18_34, Pct65Plus,
    AgeDependency, OldAgeDependency, ChildDependency,
    
    # Other (NO duplicates)
    UnemploymentRate2020, NetMigration2020,
    BirthRate2020, DeathRate2020,
    AvgWage, PropSalaries,
    NumTaxReturns, PartnershipIncome,
    NumEicReturns, ChildTaxCredit
    
    # REMOVED: PctNoHs, PctWhite (said dropped), PctMale (duplicate),
    #          NetMigration2020 (duplicate), DeathRate2020 (duplicate)
  ) %>%
  drop_na()

# Sanity check BEFORE running best subsets
cat("Response variable range:\n")
print(summary(df_model$sq_PresDemRatio))
# Should be ~0.2 to 0.9 (square roots of proportions), NOT 0 to 1

cat("Variables in model data:", ncol(df_model) - 1, "\n")
cat("Sample size:", nrow(df_model), "\n")

# ---------------- BEST SUBSETS ----------------
regfit <- regsubsets(
  sq_PresDemRatio ~ .,
  data    = df_model,
  nvmax   = 20,
  method  = "seqrep"
)

# Plot
plot(regfit, scale = "adjr2",
     main = "Best Subsets",
     cex.axis = 0.55)

# Summary table
get_best <- function(model) {
  s <- summary(model)
  data.frame(
    NumVars = 1:length(s$adjr2),
    AdjR2   = round(s$adjr2, 4)
  ) %>% arrange(desc(AdjR2))
}

best_models <- get_best(regfit)
head(best_models, 10)

s <- summary(regfit)
cat("\nBest model by Adjusted R2:", which.max(s$adjr2), "variables\n")

plot(s$adjr2, type = "b", xlab = "# Variables", ylab = "Adj R2", main = "Adjusted R2")
points(which.max(s$adjr2), max(s$adjr2), col = "red", pch = 16, cex = 1.5)

# ---------------- FIT BEST MODEL ----------------
best_n    <- which.max(s$adjr2)
coef_best <- coef(regfit, best_n)
vars_best <- names(coef_best)[-1]
cat("\nVariables in best model (Adj R2):\n")
print(vars_best)

best_formula <- as.formula(
  paste("sq_PresDemRatio ~", paste(vars_best, collapse = " + "))
)

lm_best <- lm(best_formula, data = df_model)
summary(lm_best)

# ---------------- RESIDUAL DIAGNOSTICS ----------------
par(mfrow = c(1, 1))
plot(lm_best)
par(mfrow = c(1, 1))

# ---------------- RESIDUALS VS EACH PREDICTOR ----------------
for (v in vars_best) {
  plot(df_model[[v]], lm_best$residuals,
       xlab = v,
       ylab = "Residuals",
       main = paste("Residuals vs", v),
       pch  = 16, cex = 0.5,
       col  = adjustcolor("steelblue", alpha.f = 0.5))
  abline(h = 0, col = "red", lty = 2)
}
```

