---
title: "An Extra Sum of Stress: Does Economic Distress Drives Counties to Support the Democratic Party?"
author: "Khanh Do, Maral Bat-Erdene"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggrepel)
library(readr)
library(tidyverse)
library(corrplot)
require(leaps)
library(patchwork)
require(mosaic)
require(corrplot)
require(leaps)
library(ggplot2)
library(reshape2)
library(car)
library(broom)
```

# Overview

## About our Data

```{r message = FALSE, warning = FALSE, echo = FALSE}
url <- "https://raw.githubusercontent.com/khanhdo05/sta-310-final/main/data/clean/FINAL_DATA.csv"
df <- read_csv(url, show_col_types = FALSE)
```

Read csv from this [link](https://raw.githubusercontent.com/khanhdo05/sta-310-final/main/data/clean/FINAL_DATA.csv) into a dataframe `df`.

Code book available [here](https://docs.google.com/spreadsheets/d/1A-s4d9eTGT6rGAFhPsAnOutXermwbm6LxZr9JXc_NPE/edit?gid=0#gid=0).

Each row is unique by column `FIPS`.

```{r echo=FALSE}
cat("Original data's nubmer of rows:", nrow(df), "\n")
cat("Original data's nubmer of columns:", ncol(df), "\n")
```

## Abstract

This analysis investigates whether **economic distress markers** provide meaningful explanatory power for predicting county-level Democratic presidential vote share in 2020, *beyond* what traditional demographic, education, and prosperity variables already capture.

Our economic distress block consists of seven indicators:

- **`UnemploymentRate2020`** — the raw percentage of the civilian labor force actively without work in 2020

- **`PropUnemplComp`** — unemployment compensation as a share of total gross income, measuring county-level reliance on government safety nets rather than active earnings

- **`PropEICAmount`** — Earned Income Credit (EIC) amount as a share of total gross income, measuring the intensity of low-income household prevalence. The EIC is a federal refundable tax credit targeting low-to-moderate income working individuals and families; a high `PropEICAmount` signals a concentration of working-poor households.

- **`PropChildTax`** — Additional Child Tax Credit (ACTC) amount as a share of total gross income. The ACTC is a *refundable* credit available to low-income families with children who owe little to no federal tax. Counties with high `PropChildTax` relative to income reflect a concentration of working-poor families with children, making this a legitimate economic distress marker.

- **`SeniorBurden`** — the old-age dependency ratio: the share of the population aged 65+ relative to the working-age population (18–64). Counties with a large elderly population place a higher fiscal dependency load on a shrinking working-age tax base, stressing both county-level public finances and household income composition.

- **`PropSocSec`** — Social Security benefits as a share of total gross income, serving as a proxy for elderly income dependency

- **`PropPensions`** — taxable pensions and annuities as a share of total gross income, a further indicator of retiree income reliance

The investigation is rooted in an **Extra Sum of Squares (ESS) F-test**: we first build a Reduced Model (Model R) using traditional demographic, education, and prosperity predictors, then build a Full Model (Model F) by adding the economic distress block, and formally test whether the distress block significantly reduces model error.

This approach follows Autor, Dorn, and Hanson (2020), who found that counties experiencing economic decline sometimes shift *away* from Democratic candidates, and Gelman (2008), who documented the complex relationship between income, geography, and partisan voting. We aim to determine whether localized economic stress offers measurable predictive power beyond general income levels.

# 1. Data Transformation

## 1.1 Response Variable: PresDemRatio

```{r echo=FALSE}
hist(df$PresDemRatio,
     main = "Distribution of PresDemRatio",
     col = "steelblue")
```

The data looks right-skewed, but we shouldn't use log-transformation because this variable is a proportion bounded by 0 and 1. Log-transformation is too aggressive for this data and can turn a right-skewed distribution to left-skewed. A square-root transformation is milder and more appropriate. We can see this more clearly with the histograms below.

```{r echo=FALSE}
hist(log(df$PresDemRatio),
     main = "Distribution of Log-Transformed PresDemRatio",
     col = "steelblue")
hist(sqrt(df$PresDemRatio),
     main = "Distribution of Sqrt-Transformed PresDemRatio",
     col = "steelblue")
```

The square-root transformation produces a substantially more symmetric, bell-shaped distribution and is applied to the response for all subsequent modeling.

```{r echo=FALSE}
# Apply sqrt transformation to response
df1 <- df %>%
  mutate(sq_PresDemRatio = sqrt(PresDemRatio)) %>%
  relocate(sq_PresDemRatio) %>%
  select(-PresDemRatio)
```


## 1.2 Predictor Variables

### 1.2.1 Removing Structurally Redundant Columns

Before any analysis, we remove columns that are structurally incompatible with multiple regression for purely mathematical or scope reasons. These removals are *not* conceptual decisions about the economic distress block — they are necessary preprocessing.

**Raw counts removed in favor of proportion/average counterparts** (to reduce scale-driven multicollinearity):

```{r echo=FALSE}
df2 <- df1 %>%
  select(
    # Raw dollar counts — proportion or average counterparts are retained instead
    -SalariesWages, 
    -BusinessNetIncome, 
    -CapitalGains, 
    -PensionsAnnuities, 
    -UnemploymentComp, 
    -SocialSecurityBenefits, 
    -PartnershipIncome, 
    -MortgageInterest, 
    -CharitableContributions, 
    -TotalTaxLiability, 
    -EICAmount,
    -ChildTaxCredit,
    
    # Raw headcounts — proportion counterparts are retained instead
    -NumEICReturns,
    -NetMigration2020,
    -CensusPop2020,
    -NumTaxReturns,
    
    # Perfect linear combination: SeniorAndYouthBurden = SeniorBurden + YouthBurden
    # Retaining both components and their sum would create a singular design matrix
    -SeniorAndYouthBurden,
    
    # YouthBurden: child dependency is a demographic characteristic; our distress
    # block focuses on senior/elderly fiscal dependency (SeniorBurden). Retaining
    # both YouthBurden and SeniorBurden alongside their removed sum also risks
    # near-perfect collinearity.
    -YouthBurden
  )
cat("Columns before removing structural redundancies:", ncol(df1), "\n")
cat("Columns after removing structural redundancies:", ncol(df2), "\n")
```

> **Note on `PropEICReturns`:** This variable (breadth of EIC claims — the *proportion* of filers claiming EIC) is conceptually related to economic distress and is retained at this stage. Its disposition relative to `PropEICAmount` (intensity of EIC claims) is addressed explicitly when we define the economic distress block in Section 1.2.4.

### 1.2.2 Missing Data

```{r missing-data, echo=FALSE}
missing_summary <- df2 %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing") %>%
  mutate(
    Total       = nrow(df2),
    Pct_Missing = round(100 * Missing / Total, 1)
  ) %>%
  arrange(desc(Missing))

cat("Maximum % missing across all variables:",
    max(missing_summary$Pct_Missing), "%\n")
cat("Variables with any missing values:\n")
print(missing_summary %>% filter(Missing > 0))
```

No variable exceeds 0.5% missing. We proceed by removing rows with any `NA` values.

```{r echo=FALSE}
# remove rows with NAs
df3 <- df2 %>%
  drop_na()

cat("Rows before removing NAs:", nrow(df2), "\n")
cat("Rows after removing NAs: ", nrow(df3), "\n")

# Track sample size
n_original  <- nrow(df)
n_after_na  <- nrow(df3)
```

### 1.2.3 Outlier Detection

**Small Population:**

Two counties with fewer than 100 residents — Loving County, TX (pop. 64) and Kalawao County, HI (pop. 82) — are removed prior to any modeling. With such small populations, all derived income proportion variables are computed from an extremely small base of tax filers, making them unreliable estimates of county-level economic conditions. This decision requires no model information and is made on data quality grounds alone.

```{r echo=FALSE}
small_counties <- df %>% 
  filter(CensusPop2020 < 100) %>% 
  select(FIPS, CountyName, CensusPop2020)

cat("Counties removed (population < 100):\n")
print(small_counties)

df3 <- df3 %>% filter(!FIPS %in% small_counties$FIPS)
n_after_small <- nrow(df2)
```

**Cook's Distance:**

Cook's Distance (Cook, 1977) is used as the primary outlier detection method because it measures the influence of each observation on the fitted regression coefficients — directly relevant to our regression modeling goal. Unlike univariate methods (IQR, Z-score) that flag observations based on a single variable's distribution alone, Cook's D identifies observations whose removal would substantially change the estimated predictor–response relationship (Belsley, Kuh & Welsch, 1980). The threshold of $4/n$ (Cook & Weisberg, 1982) is applied, where $n$ is the number of complete observations. Outliers are flagged here for awareness; removal decisions are revisited after model fitting in Section 3.

```{r echo=FALSE}
# Prep data
df_out <- df3 %>%
  select(-FIPS, -CountyName) %>%
  drop_na()

response   <- "sq_PresDemRatio"
predictors <- setdiff(names(df_out), response)

# Create a list to store plots
plot_list <- list()

# Outlier detection loop
for (var in predictors) {
  
  temp <- df_out %>%
    select(all_of(c(response, var))) %>%
    rename(x = all_of(var), y = all_of(response))
  
  n <- nrow(temp)
  model   <- lm(y ~ x, data = temp)
  cooks_d <- cooks.distance(model)
  cutoff  <- 4 / n
  
  temp <- temp %>%
    mutate(cooks_d = cooks_d, is_outlier = cooks_d > cutoff)
  
  # Create the ggplot and save it to the list
  p <- ggplot(temp, aes(x = x, y = y)) +
    geom_point(aes(color = is_outlier), alpha = 0.4) +
    scale_color_manual(values = c("gray60", "red")) +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    labs(
      title = paste(var),
      subtitle = paste("Outliers:", sum(temp$is_outlier)),
      x = NULL, y = NULL
    ) +
    theme_minimal(base_size = 8) +
    theme(legend.position = "none")
  
  plot_list[[var]] <- p
}

# Display plots
wrap_plots(plot_list[1:9], ncol = 3, nrow = 3) +
  plot_annotation(title = "Outlier Detection 1")
wrap_plots(plot_list[10:18], ncol = 3, nrow = 3) +
  plot_annotation(title = "Outlier Detection 2")
wrap_plots(plot_list[19:27], ncol = 3, nrow = 3) +
  plot_annotation(title = "Outlier Detection 3")
wrap_plots(plot_list[28:35], ncol = 3, nrow = 3) +
  plot_annotation(title = "Outlier Detection 4")
```

### 1.2.4 Correlation Structure and Transformation Decisions

We examine the correlation matrix and pairwise scatterplots by thematic group to guide transformation choices and identify multicollinearity clusters.

```{r echo=FALSE}
# Remove categorical/binary variables before computing correlation
df_corr <- df_out %>%
  select(-IsMetro) %>%
  select(where(is.numeric))

corrplot(cor(df_corr, use = "complete.obs"),
         title   = "Correlation Matrix — All Numeric Predictors",
         mar     = c(0, 0, 1, 0),
         tl.cex  = 0.5)
```

**Key observations:**

- **Positive correlates of Democratic vote share:** education (`PctBachelors`), racial diversity (`PctBlack`, `PctAsian`, `PctHispanic`), urbanization (`PctUrban`), average wages (`AvgWage`), mortgage reliance (`PropMortgage`), charitable giving (`PropCharity`), and tax liability proportion (`PropTaxLiability`).

- **Negative correlates:** white population share (`PctWhite`), high school-only education (`PctHSOnly`), senior dependency (`SeniorBurden`), and Social Security reliance (`PropSocSec`).

- **Distress markers and voting:** Several economic distress indicators — including `PropEICAmount` — show a *negative* correlation with Democratic vote share. This is consistent with Autor et al. (2020), who found that economic decline in certain communities correlates with movement *away* from Democratic candidates rather than toward them. This is precisely why an ESS test is warranted: the relationship is not straightforward.

**Multicollinearity:** The matrix highlights several clusters where variables are so highly correlated that they are likely redundant:

- **Economic distress:** There are very large blue circles between PropEICAmount, PropEICReturns and PropChildTax. SeniorBurden is also highly correlated with Pct65Plus and MedianAge so possible interaction might be needed in the full model. 

- **Education:** PctBachelors is strongly negatively correlated with PctNoHS and PctHSOnly. We will likely pick only one (PctBachelors) to represent education levels in our Reduced Model to avoid multicollinearity.

- **Age:** There is extreme correlation between MedianAge and Pct65Plus. We will use Best Subsets Regression later to determine which single age-related variable is the most efficient predictor.

- **Race:** PctWhite has prominent red circles at its intersections with PctBlack, PctHispanic, and PctAsian. This is expected in a closed system where these variables represent parts of a whole (100% of the population). We will Likely select PctWhite (as a baseline) and PctBlack and PctAsian (as markers of diversity)—rather than using the entire set.

- **Prosperity:** TotalGrossIncome shows a positive relationship with AvgWage and PropCharity, but its correlation with the vote share is less pronounced than the "per-capita" or "proportion" versions. PropBusiness and AvgBusiness have strong correlation as well since they share the same numerator - we will likely just use PropBusiness because it aligns with our research into general prosperity.

**Decision on `PropEICReturns` vs. `PropEICAmount`:**
Both measure different dimensions of EIC-related economic distress: `PropEICReturns` captures the *breadth* of poverty (what share of filers claim EIC), while `PropEICAmount` captures the *intensity* (how much EIC is claimed relative to total income). Their correlation is `r round(cor(df3$PropEICReturns, df3$PropEICAmount, use = "complete.obs"), 3)`, indicating near-redundancy. For the economic distress block, we retain `PropEICAmount` (intensity) as the more theoretically meaningful indicator of the *depth* of financial vulnerability in a county, and exclude `PropEICReturns`.

```{r drop-eic-returns, echo=FALSE}
df3 <- df3 %>% select(-PropEICReturns)
```

#### Group-by-Group Pairwise Plots and Transformation Rationale

**Group 1: General Prosperity and Affluence**

```{r echo=FALSE}
# Group 1: General Prosperity and Affluence
pairs(df3 %>% select(sq_PresDemRatio, TotalGrossIncome, PropCapitalGains, PropPartnership, 
                         PropMortgage, PropCharity),
      main = "General Prosperity vs Voting 1",
      pch = 16, cex = 0.3,
      col = adjustcolor("steelblue", alpha.f = 0.4))
```

```{r echo=FALSE}
# Group 1: General Prosperity and Affluence
pairs(df3 %>% select(sq_PresDemRatio, PropTaxLiability, PropSalaries, AvgWage, PropBusiness, AvgBusiness),
      main = "General Prosperity vs Voting 2",
      pch = 16, cex = 0.3,
      col = adjustcolor("steelblue", alpha.f = 0.4))
```

Raw income totals (`TotalGrossIncome`) are heavily right-skewed and span multiple orders of magnitude → **log transformation**. 

`PropCapitalGains` is also right-skewed and strictly non-negative → **log transformation** (after verifying no zero/negative values). 

`PropCharity` shows a right-skewed distribution with a heavy upper tail → examine more carefully below before deciding. 

`PropBusiness` and `AvgBusiness` can contain negative values (business losses), ruling out log transformation → **Inverse Hyperbolic Sine (IHS) transformation** (`asinh`) handles negative values while approximating log-scale compression for large positive values. We retain `PropBusiness` (relative to income) over `AvgBusiness` (per filer) as it better aligns with our comparative framework.

Other ratios (`PropSalaries`, `PropTaxLiability`) are mildly skewed proportions → **square-root transformation**.

```{r check-propcapgains, echo=FALSE}
cat("Min PropCapitalGains:", min(df3$PropCapitalGains, na.rm = TRUE), "\n")

# Visualize PropCharity vs response before deciding
ggplot(df3, aes(x = PropCharity, y = sq_PresDemRatio)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "loess", color = "red") +
  geom_smooth(method = "lm", color = "blue", linetype = "dashed") +
  labs(title = "PropCharity vs. sq_PresDemRatio (Raw)",
       subtitle = "LOESS (red) vs. Linear fit (blue dashed)") +
  theme_minimal()

ggplot(df3, aes(x = sqrt(PropCharity), y = sq_PresDemRatio)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "loess", color = "red") +
  geom_smooth(method = "lm", color = "blue", linetype = "dashed") +
  labs(title = "sqrt(PropCharity) vs. sq_PresDemRatio",
       subtitle = "LOESS (red) vs. Linear fit (blue dashed)") +
  theme_minimal()
```

**Group 2: Economic Distress and Financial Vulnerability**

```{r echo=FALSE}
# Group 2: Economic Distress and Financial Vulnerability
pairs(df3 %>% select(sq_PresDemRatio, UnemploymentRate2020, PropUnemplComp, 
                     PropSocSec, PropPensions, PropEICAmount, PropChildTax, 
                     SeniorBurden),
      main = "Economic Distress vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("firebrick", alpha.f = 0.4))
```

Many distress markers show a *negative* association with Democratic vote share — counties with more severe distress tend to vote less Democratic, consistent with Autor et al. (2020). All variables here are non-negative ratios or rates with mild-to-moderate right skew → **square-root transformation**. Note that `PropEICAmount` and `PropChildTax` show strong internal correlation, as both capture financial vulnerability of low-income families; they are retained as distinct distress dimensions. 

**Group 3: Educational Attainment**

```{r echo=FALSE}
# Group 3: Educational Attainment
pairs(df3 %>% select(sq_PresDemRatio, PctNoHS, PctHSOnly, PctSomeCollege, PctBachelors),
      main = "Education Levels vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("darkgreen", alpha.f = 0.4))
```

`PctBachelors` is the strongest positive predictor of Democratic vote share. Strong negative correlations between `PctBachelors` and lower education categories indicate multicollinearity; we will select `PctBachelors` as the single education representative in the Reduced Model. All are percentage variables → **square-root transformation** except for `PctSomeCollege` which looks good already.

**Group 4: Racial and Gender Demographics**

```{r echo=FALSE}
# Group 4: Racial and Gender Demographics
pairs(df3 %>% select(sq_PresDemRatio, PctWhite, PctBlack, PctAsian, PctHispanic, PctMale),
      main = "Demographics vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("darkorchid", alpha.f = 0.4))
```

Race variables sum to approximately 100%, creating zero-sum multicollinearity. We check whether `PctBlack` has a quadratic relationship with the response (given its right-skewed distribution in many counties):

```{r check-pctblack-quadratic, echo=FALSE}
ggplot(df3, aes(x = sqrt(PctBlack), y = sq_PresDemRatio)) +
  geom_point(alpha = 0.3, color = "darkorchid") +
  geom_smooth(method = "loess", color = "red") +
  geom_smooth(method = "lm", color = "blue", linetype = "dashed") +
  labs(title = "sqrt(PctBlack) vs. sq_PresDemRatio",
       subtitle = "LOESS (red) vs. Linear fit (blue dashed) — checking for quadratic pattern") +
  theme_minimal()

# Also check quadratic fit
lm_black_lin  <- lm(sq_PresDemRatio ~ sqrt(PctBlack), data = df3)
lm_black_quad <- lm(sq_PresDemRatio ~ sqrt(PctBlack) + I(sqrt(PctBlack)^2), data = df3)
cat("Linear R²:   ", round(summary(lm_black_lin)$r.squared, 4), "\n")
cat("Quadratic R²:", round(summary(lm_black_quad)$r.squared, 4), "\n")
cat("Quadratic term p-value:",
    round(summary(lm_black_quad)$coefficients[3, 4], 4), "\n")
```

`PctMale` is a percentage variable near 50% with low skewness, but we apply a log transformation per reviewer recommendation to stabilize variance in the tails. All other race variables → **square-root transformation**.

**Group 5: Age and Dependency Ratios**

```{r echo=FALSE}
# Group 5: Age and Dependency Ratios
pairs(df3 %>% select(sq_PresDemRatio, MedianAge, Pct18_34, Pct65Plus, SeniorBurden),
      main = "Age and Dependency vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("darkorange", alpha.f = 0.4))
```

`MedianAge`, `Pct65Plus`, and `SeniorBurden` are highly redundant. `SeniorBurden` is retained as it directly captures the fiscal dependency ratio relevant to our distress block. → **square-root transformation** for the dependency ratio. Note that `SeniorBurden` is part of the distress block and will not be in the reduced model.

**Group 6: Urbanization and Population Dynamics**

```{r echo=FALSE}
# Group 6: Urbanization and Population Dynamics
pairs(df3 %>% select(sq_PresDemRatio, PctUrban, IsMetro, BirthRate2020, 
                         DeathRate2020, MigrationRate2020),
      main = "Urbanization and Dynamics vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("dodgerblue", alpha.f = 0.4))
```

This group captures the urban-rural divide, with metropolitan integration (IsMetro) acting as a clear categorical separator. `PctUrban` shows a non-linear positive relationship with Democratic voting, capturing the urban–rural divide. → **square-root transformation**. `BirthRate2020` and `DeathRate2020` are rates → **square-root transformation** to adress mild skewness.

### 1.2.5 Best Subsets (Before predictor transformations)

```{r}
df_reduced_candidates <- df3 %>%
  select(
    -UnemploymentRate2020,
    -PropUnemplComp,
    -PropEICAmount,
    -SeniorBurden,
    -PropChildTax,
    -PropSocSec,
    -PropPensions
  )

fit_g1 <- regsubsets(
  sq_PresDemRatio ~ . -FIPS -CountyName,
  data   = df_reduced_candidates,
  nvmax  = 20,
  method = "seqrep"
)

s_g1 <- summary(fit_g1)

best_models_g1 <- data.frame(
  NumVars = 1:length(s_g1$adjr2),
  AdjR2   = round(s_g1$adjr2, 4),
  s_g1$which
) %>% arrange(desc(AdjR2))

# Plot
plot(s_g1$adjr2, type = "b",
     xlab = "Number of Variables",
     ylab = "Adjusted R2")
best_n_g1 <- which.max(s_g1$adjr2)
points(best_n_g1, max(s_g1$adjr2), col = "red", pch = 16, cex = 1.5)

# Extract the which matrix and adjr2
mat    <- s_g1$which          # logical matrix: rows = model size, cols = variables
adjr2  <- s_g1$adjr2          # adj R² for each model size

# Build a data frame (drop the intercept column)
df_plot        <- as.data.frame(mat[, -1])   # remove "(Intercept)" column
df_plot$adjr2  <- round(adjr2, 4)
df_plot$model  <- seq_len(nrow(df_plot))

# Melt to long format
df_long <- melt(
  df_plot,
  id.vars       = c("model", "adjr2"),
  variable.name = "Variable",
  value.name    = "Included"
)

# Plot
ggplot(df_long, aes(x = Variable, y = factor(adjr2), fill = Included)) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c("TRUE" = "black", "FALSE" = "white")) +
  labs(
    title = "Variables Selection for Reduced Model (Best Subsets)",
    x     = NULL,
    y     = "adj R²"
  ) +
  theme_minimal() +
  theme(
    axis.text.x    = element_text(angle = 45, hjust = 1, vjust = 1, size = 9),
    legend.position = "none",
    panel.grid     = element_blank()
  )
```


### 1.2.6 Applying Transformations

```{r echo=FALSE}
# ---------------- APPLY TRANSFORMATIONS ----------------
df4 <- df3 %>%
  mutate(
    # group 1: General Prosperity
    log_TotalGrossIncome  = log1p(TotalGrossIncome),
    log_PropCapitalGains  = log(PropCapitalGains + 1e-6),  # add small constant to handle zeros
    sq_PropSalaries       = sqrt(PropSalaries),
    sq_PropCharity        = sqrt(PropCharity),
    sq_PropTaxLiability   = sqrt(PropTaxLiability),
    ihs_PropBusiness = asinh(PropBusiness),
    
    # group 2: Economic Distress
    sq_PropUnemplComp     = sqrt(PropUnemplComp),
    sq_PropEICAmount      = sqrt(PropEICAmount),
    sq_PropChildTax       = sqrt(PropChildTax),
    sq_PropSocSec         = sqrt(PropSocSec),
    sq_PropPensions       = sqrt(PropPensions),
    sq_SeniorBurden       = sqrt(SeniorBurden),
    
    # group 3: Education
    sq_PctBachelors = sqrt(PctBachelors),
    sq_PctNoHS = sqrt(PctNoHS),
    sq_PctHSOnly = sqrt(PctHSOnly),
    
    # group 4: Race and Gender
    sq_PctWhite           = sqrt(PctWhite),
    sq_PctBlack           = sqrt(PctBlack),
    sq_PctHispanic        = sqrt(PctHispanic),
    sq_PctAsian           = sqrt(PctAsian),
    log_PctMale           = log(PctMale),
    
    # group 5: Age 
    # None, SeniorBurden was handled in group 2
    
    # group 6: Urbanization
    sq_PctUrban = sqrt(PctUrban),
    sq_BirthRate = sqrt(BirthRate2020),
    sq_DeathRate = sqrt(DeathRate2020),
    
  ) %>%
  select(
    # Drop untransformed originals
    -TotalGrossIncome, -PropCapitalGains, -PropSalaries, -PropCharity,
    -PropTaxLiability,
    -PropUnemplComp, -PropEICAmount, -PropChildTax, -PropSocSec, -PropPensions,
    -PctBachelors, -PctNoHS, -PctHSOnly,
    -PctWhite, -PctBlack, -PctHispanic, -PctAsian, -PctMale,
    -SeniorBurden,
    -PctUrban, -BirthRate2020, -DeathRate2020,
    -PropBusiness, -AvgBusiness  # AvgBusiness dropped in favor of PropBusiness
  )

# ---------------- VERIFY TRANSFORMATIONS ----------------
par(mfrow = c(6, 4), mar = c(2, 2, 2, 1))
transformed_vars <- c(
  "log_TotalGrossIncome", "log_PropCapitalGains", "sq_PropSalaries",
  "sq_PropCharity", "sq_PropTaxLiability",
  "sq_PropUnemplComp", "sq_PropEICAmount", "sq_PropChildTax",
  "sq_PropSocSec", "sq_PropPensions",
  "sq_PctBachelors", "sq_PctNoHS", "sq_PctHSOnly",
  "sq_PctWhite", "sq_PctBlack", "sq_PctHispanic", "sq_PctAsian", "log_PctMale",
  "sq_SeniorBurden",
  "sq_PctUrban", "sq_BirthRate", "sq_DeathRate",
  "ihs_PropBusiness"
)

for (v in transformed_vars) {
  # Verify variable exists in df4 to avoid errors
  if (v %in% names(df4)) {
    hist(df4[[v]], 
         main = v, 
         xlab = "", 
         col = "steelblue", 
         border = "white",
         cex.main = 0.8) # Smaller title text to fit the grid
  } else {
    message(paste("Warning: Variable", v, "not found in df4."))
  }
}
par(mfrow = c(1, 1))

cat("Number of columns now:", ncol(df2), "\n")
cat("Number of rows now:", nrow(df2), "\n")
```


**Summary of transformation outcomes:**

- `log_TotalGrossIncome` and `log_PropCapitalGains` are now roughly symmetric.

- Distress markers (`sq_PropUnemplComp`, `sq_PropEICAmount`, `sq_PropChildTax`) are no longer heavily clustered at zero, allowing the model to more accurately estimate the impact of financial vulnerability.

- `sq_PctBlack`, `sq_PctAsian`, and `sq_PctHispanic` remain right-skewed due to the many U.S. counties with very small minority populations — this is an inherent feature of the data, not a transformation failure.

- `sq_PctUrban` retains a spike near zero (entirely rural counties) but is more spread out than the raw variable.

- `ihs_PropBusiness` is unimodal and handles negative values (business losses) successfully.

### 1.2.6 Mean-centering

All numeric predictors are mean-centered (subtracting their mean). This is essential before including interaction terms to reduce multicollinearity between main effects and their products, and to make the intercept interpretable as the predicted response at average predictor values. `FIPS`, `IsMetro`, and the response `sq_PresDemRatio` are excluded.

```{r echo=FALSE}
df5 <- df4 %>%
  mutate(across(where(is.numeric) & !any_of(c("FIPS", "sq_PresDemRatio", "IsMetro")), 
                ~ . - mean(., na.rm = TRUE)))

col_means <- df5 %>%
  select(where(is.numeric) & !any_of(c("FIPS", "sq_PresDemRatio", "IsMetro"))) %>%
  summarise(across(everything(), ~ round(mean(., na.rm = TRUE), 6)))
cat("Column means after centering (should all be ~0):\n")
print(t(col_means))
```

---

# 2. Best Subsets After Predictor Transformation

We will now run Best Subsets Regression on our dataset (excluding the economic distress block) to select variables for our Reduced Model.

```{r echo=FALSE}
df_reduced_candidates <- df5 %>%
  select(
    -UnemploymentRate2020,
    -sq_PropUnemplComp,
    -sq_PropEICAmount,
    -sq_SeniorBurden,
    -sq_PropChildTax,
    -sq_PropSocSec,
    -sq_PropPensions
  )

fit_g1 <- regsubsets(
  sq_PresDemRatio ~ . -FIPS -CountyName,
  data   = df_reduced_candidates,
  nvmax  = 20,
  method = "seqrep"
)

s_g1 <- summary(fit_g1)

best_models_g1 <- data.frame(
  NumVars = 1:length(s_g1$adjr2),
  AdjR2   = round(s_g1$adjr2, 4),
  s_g1$which
) %>% arrange(desc(AdjR2))

# Plot
plot(s_g1$adjr2, type = "b",
     xlab = "Number of Variables",
     ylab = "Adjusted R2")
best_n_g1 <- which.max(s_g1$adjr2)
points(best_n_g1, max(s_g1$adjr2), col = "red", pch = 16, cex = 1.5)

# Extract the which matrix and adjr2
mat   <- s_g1$which
adjr2 <- s_g1$adjr2

# Build data frame (drop intercept column)
df_plot       <- as.data.frame(mat[, -1])
df_plot$adjr2 <- round(adjr2, 4)
df_plot$model <- seq_len(nrow(df_plot))

# Melt to long format
df_long <- melt(
  df_plot,
  id.vars       = c("model", "adjr2"),
  variable.name = "Variable",
  value.name    = "Included"
)

# Plot
ggplot(df_long, aes(x = Variable, y = factor(adjr2), fill = Included)) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c("TRUE" = "black", "FALSE" = "white")) +
  labs(
    title = "Group 1 — Variables Selection for Reduced Model (Best Subsets)",
    x     = NULL,
    y     = "adj R²"
  ) +
  theme_minimal() +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, vjust = 1, size = 9),
    legend.position = "none",
    panel.grid      = element_blank()
  )
```

The transformed variables not only improve model assumptions but also improved overall $R^2$.

---

# 3. Reduced Model

The **Reduced Model (Model R)** uses only the predictors selected by Best Subsets — traditional demographic, education, prosperity, and urbanization variables. It deliberately excludes the economic distress block. This is the baseline against which the Full Model will be compared in the Extra Sum of Squares test. 

## 3.1 Reduced Model Without Interactions

```{r echo=FALSE}
cat("--- REDUCED MODEL WITHOUT INTERACTION SUMMARY --- \n")
cat("Best model size (Adj R2):", best_n_g1, "variables\n")
cat("Best Adj R2:", round(max(s_g1$adjr2), 4), "\n")
# Selected variables
vars_g1 <- names(coef(fit_g1, best_n_g1))[-1]
cat("Selected from the data:\n")
print(vars_g1)

# Build formula dynamically from Best Subsets selected variables
formula_reduced1 <- as.formula(
  paste("sq_PresDemRatio ~", paste(vars_g1, collapse = " + "))
)

lm_reduced1 <- lm(formula_reduced1, data = df5)

par(mfrow = c(2, 2))
plot(lm_reduced1, main = "Reduced Model (No Interactions)")
par(mfrow = c(1, 1))

# ---------------- RESIDUALS VS EACH PREDICTOR ----------------
model_df <- model.frame(lm_reduced1)
res <- residuals(lm_reduced1)
par(mfrow = c(ceiling(length(vars_g1) / 4), 4), mar = c(3, 3, 2, 1))
for (v in vars_g1) {
  plot(model_df[[v]], res,
       xlab = v,
       ylab = "Residuals",
       main = paste("Residuals vs", v),
       pch  = 16, cex = 0.5,
       col  = adjustcolor("steelblue", alpha.f = 0.5))
  abline(h = 0, col = "red", lty = 2)
}
par(mfrow = c(1, 1))
```

In the **Residuals vs Fitted plot**, observation 1691 is an extreme outlier. Let's investigate this.

```{r echo=FALSE}
df5[1691,]
df %>% filter(FIPS == 31149)
```

Rock County has `SalariesWages` $>$ `TotalGrossIncome`. Let's check all counties that are similar.

```{r echo=FALSE}
# Flag counties where salaries exceed total gross income (data anomaly)
anomalies <- df %>% 
  filter(SalariesWages > TotalGrossIncome) %>%
  select(FIPS, CountyName, NumTaxReturns, TotalGrossIncome, SalariesWages)

cat("Counties where SalariesWages > TotalGrossIncome:\n")
print(anomalies)
```

Four counties have this same data anomaly — salaries exceeding total gross income. Thus, an additional four counties were removed on data quality grounds: Hayes County, NE; Rock County, NE; Sioux County, ND; and Shackelford County, TX. In each case, reported salaries and wages exceeded total adjusted gross income — a mathematical inconsistency arising when large business or partnership losses reduce AGI below gross wages. This renders all income proportion variables for these counties unreliable (several exceed 1.0) and they are excluded from analysis. 

```{r echo=FALSE}
anomaly_fips <- c(31085, 31149, 38085, 48417)

anomaly_counties <- df %>%
  filter(FIPS %in% anomaly_fips) %>%
  select(FIPS, CountyName, NumTaxReturns, TotalGrossIncome, SalariesWages)

df6 <- df5 %>%
  filter(!FIPS %in% anomaly_fips)

cat("\nObservations removed:", nrow(df5) - nrow(df6), "\n")
cat("Final analytic sample:", nrow(df6), "counties\n")
```

Now, we rerun Best Subsets since we removed four more observations.

```{r echo=FALSE}
df_reduced_candidates <- df6 %>%
  select(
    -UnemploymentRate2020,
    -sq_PropUnemplComp,
    -sq_PropEICAmount,
    -sq_SeniorBurden,
    -sq_PropChildTax,
    -sq_PropSocSec,
    -sq_PropPensions
  )

fit_g2 <- regsubsets(
  sq_PresDemRatio ~ . -FIPS -CountyName,
  data   = df_reduced_candidates,
  nvmax  = 20,
  method = "seqrep"
)

s_g2 <- summary(fit_g2)

best_models_g2 <- data.frame(
  NumVars = 1:length(s_g2$adjr2),
  AdjR2   = round(s_g2$adjr2, 4),
  s_g2$which
) %>% arrange(desc(AdjR2))

# Plot
plot(s_g2$adjr2, type = "b",
     xlab = "Number of Variables",
     ylab = "Adjusted R2")
best_n_g2 <- which.max(s_g2$adjr2)
points(best_n_g2, max(s_g2$adjr2), col = "red", pch = 16, cex = 1.5)

# Extract the which matrix and adjr2
mat   <- s_g2$which
adjr2 <- s_g2$adjr2

# Build data frame (drop intercept column)
df_plot       <- as.data.frame(mat[, -1])
df_plot$adjr2 <- round(adjr2, 4)
df_plot$model <- seq_len(nrow(df_plot))

# Melt to long format
df_long <- melt(
  df_plot,
  id.vars       = c("model", "adjr2"),
  variable.name = "Variable",
  value.name    = "Included"
)

# Plot
ggplot(df_long, aes(x = Variable, y = factor(adjr2), fill = Included)) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c("TRUE" = "black", "FALSE" = "white")) +
  labs(
    title = "Variables Selection for Reduced Model (Best Subsets)",
    x     = NULL,
    y     = "adj R²"
  ) +
  theme_minimal() +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, vjust = 1, size = 9),
    legend.position = "none",
    panel.grid      = element_blank()
  )
```

```{r}
cat("--- REDUCED MODEL AFTER OUTLIERS REMOVAL SUMMARY --- \n")
cat("Best model size (Adj R2):", best_n_g2, "variables\n")
cat("Best Adj R2:", round(max(s_g2$adjr2), 4), "\n")
# Selected variables
vars_g2 <- names(coef(fit_g2, best_n_g2))[-1]
cat("Selected from the data:\n")
print(vars_g2)

# Build formula dynamically from Best Subsets selected variables
formula_reduced2 <- as.formula(
  paste("sq_PresDemRatio ~", paste(vars_g2, collapse = " + "))
)

lm_reduced2 <- lm(formula_reduced2, data = df6)

par(mfrow = c(2, 2))
plot(lm_reduced2, main = "Reduced Model (No Interactions)")
par(mfrow = c(1, 1))
```

The $R^2$ is not overwhelmingly better but the residual plots look better.

## 3.2 Identifying Interaction Terms

We will now test which interaction terms can be important.

```{r}
# Build saturated two-way interaction formula from the same variables
formula_interactions <- as.formula(
  paste("sq_PresDemRatio ~ (", paste(vars_g2, collapse = " + "), ")^2")
)

# Fit interaction model
lm_interactions <- lm(formula_interactions, data = df6)

tidy_int <- tidy(lm_interactions)

# Bonferroni-corrected threshold
n_tests <- sum(grepl(":", tidy_int$term))
bonf_alpha <- 0.05 / n_tests
cat("Number of interaction terms tested:", n_tests, "\n")
cat("Bonferroni-corrected alpha:", round(bonf_alpha, 6), "\n")

top_interactions <- tidy_int %>%
  filter(grepl(":", term)) %>%
  filter(p.value < bonf_alpha) %>%
  arrange(p.value) %>%
  slice(1:10)

cat("\nStatistically significant interactions (Bonferroni-corrected):\n")
print(top_interactions %>% select(term, estimate, std.error, statistic, p.value))

significant_main <- tidy_int %>%
  filter(!grepl(":", term), term != "(Intercept)", p.value < 0.05) %>%
  arrange(p.value)

cat("\nSignificant main effects:\n")
print(significant_main %>% select(term, estimate, p.value))
```

## 3.3 Final Reduced Model

We build the final Reduced Model using the selected main effects and theoretically motivated interaction terms that also reach statistical significance.

```{r echo=FALSE}
selected_terms <- c(
  significant_main$term,
  top_interactions$term
)

formula_reduced <- as.formula(
  paste("sq_PresDemRatio ~", paste(selected_terms, collapse = " + "))
)

lm_R <- lm(formula_reduced, data = df6)

cat("--- FINAL REDUCED MODEL (Model R) SUMMARY ---\n")
summary(lm_R)

par(mfrow = c(2, 2))
plot(lm_R, main = "Model R Diagnostics")
par(mfrow = c(1, 1))

# ---------------- RESIDUALS VS EACH PREDICTOR ----------------
mf_final  <- model.frame(lm_R)
res_final <- residuals(lm_R)

# Split main effects and interactions
main_terms <- selected_terms[!grepl(":", selected_terms)]
inter_terms <- selected_terms[grepl(":", selected_terms)]

# Build values for all terms
term_vals <- lapply(selected_terms, function(term) {
  if (grepl(":", term)) {
    parts <- strsplit(term, ":")[[1]]
    mf_final[[parts[1]]] * mf_final[[parts[2]]]
  } else {
    mf_final[[term]]
  }
})
names(term_vals) <- selected_terms

par(mfrow = c(ceiling(12 / 4), 4), mar = c(3, 3, 2, 1))
for (v in selected_terms) {
  plot(term_vals[[v]], res_final,
       xlab = v, ylab = "Residuals",
       main = paste("Residuals vs", v),
       pch = 16, cex = 0.5, cex.main = 0.6,
       col = adjustcolor("steelblue", alpha.f = 0.5))
  abline(h = 0, col = "red", lty = 2)
}
par(mfrow = c(1, 1))
```

---

# 4. Full Model

The **Full Model** adds the seven economic distress block variables to Model R. This model represents the hypothesis that economic distress provides additional explanatory power beyond traditional predictors.

**Economic Distress Block (added to Reduced Model):**

| Variable | Description |
|---|---|
| `UnemploymentRate2020` | Raw unemployment rate — active joblessness |
| `sq_PropUnemplComp` | Reliance on unemployment compensation (safety net dependency) |
| `sq_PropEICAmount` | Intensity of EIC claims (working-poor concentration) |
| `sq_PropChildTax` | ACTC reliance (working-poor families with children) |
| `sq_SeniorBurden` | Old-age dependency ratio (fiscal stress from elderly population) |
| `sq_PropSocSec` | Social Security income reliance |
| `sq_PropPensions` | Pension and annuity income reliance |

**Economic Distress Interaction Terms (add if present in the Reduced Model):**

we consider whether any economic distress markers have theoretically motivated interactions with variables already in Reduced Model. Including these interactions in Full Model — rather than only distress main effects — allows the ESS test to capture the full contribution of economic distress as a construct, including the ways distress effects are *moderated* by demographic context.

**`sq_SeniorBurden:sq_PropSocSec` — Compounded elderly fiscal stress**

Senior burden (the old-age dependency ratio) and Social Security reliance measure related but distinct dimensions of elderly economic dependency. Their interaction captures counties where both are simultaneously high — a compounded fiscal stress condition where a large elderly population is also heavily dependent on federal transfer income rather than private wealth. The political effect of senior burden is expected to be amplified in these counties, as both the size and the economic
vulnerability of the elderly population reinforce each other.

**`sq_PropEICAmount:sq_PctBlack` and `sq_PropEICAmount:sq_PctHispanic` — Racial context of working poverty**

EIC intensity measures the concentration of working-poor households, but the political meaning of that poverty differs by racial composition. Consistent with Autor et al. (2020), high EIC concentration in predominantly white counties often correlates with a shift *away* from Democratic candidates. In counties with larger Black or Hispanic populations, the same poverty concentration is more likely to coincide with stronger Democratic turnout infrastructure and community organizing. These interactions allow the model to estimate separate slopes for EIC intensity depending on racial context, rather than forcing a single average effect across very different communities.

**`sq_PropChildTax:sq_PctHispanic` — Child poverty in Hispanic communities**

Additional Child Tax Credit (ACTC) reliance — our marker for working-poor families with children — is disproportionately concentrated in Hispanic communities nationally. This interaction tests whether the political signal of child-family poverty differs in Hispanic-majority counties relative to the baseline pattern, which would be obscured if only main effects were included.

**`sq_SeniorBurden:sq_PctBachelors` — Education moderates elderly dependency**

The political consequences of a high elderly dependency burden are likely moderated by county education levels. In highly educated counties, high senior burden may coexist with Democratic-leaning retirees and professional caregiving economies. In low-education counties, high senior burden tends to accompany resource scarcity and a more Republican-leaning older electorate. Without this interaction, the model forces a single slope for senior burden regardless of educational context.

**`sq_PropUnemplComp:MigrationRate2020` — Acute distress and population flight**

Counties with both high unemployment compensation reliance *and* high out-migration represent the most acute form of economic distress — not only are residents unemployed and reliant on safety nets, but the working-age population is actively leaving. This compounded condition of stagnation and depopulation is theoretically distinct from either variable alone and may carry a stronger or different political signal than unemployment reliance in a stable or growing county.

```{r echo=FALSE}
distress_terms <- c(
  "UnemploymentRate2020",
  "sq_PropUnemplComp",
  "sq_PropEICAmount",
  "sq_PropChildTax",
  "sq_SeniorBurden",
  "sq_PropSocSec",
  "sq_PropPensions"
)

distress_interactions <- c(
  "sq_SeniorBurden:sq_PropSocSec",
  "sq_PropEICAmount:sq_PctBlack",
  "sq_PropEICAmount:sq_PctHispanic",
  "sq_PropChildTax:sq_PctHispanic",
  "sq_SeniorBurden:sq_PctBachelors",
  "sq_PropUnemplComp:MigrationRate2020"
)

all_vars_in_data <- names(df6)
for (term in distress_interactions) {
  parts <- strsplit(term, ":")[[1]]
  present <- all(parts %in% all_vars_in_data)
  cat(term, "— both vars present:", present, "\n")
}

all_terms <- c(selected_terms, distress_terms, distress_interactions)

formula_F <- as.formula(
  paste("sq_PresDemRatio ~",
        paste(all_terms, collapse = " + "))
)

lm_F <- lm(formula_F, data = df6)

cat("--- FULL MODEL (Model F) SUMMARY ---\n")
summary(lm_F)

par(mfrow = c(2, 2))
plot(lm_F, main = "Model F — Diagnostics")
par(mfrow = c(1, 1))

cat("\nFull Model — Adjusted R²:", round(summary(lm_F)$adj.r.squared, 4), "\n")
cat("Full Model — Residual SE:", round(summary(lm_F)$sigma, 4), "\n")

# ---------------- RESIDUALS VS EACH PREDICTOR ----------------
mf_final  <- model.frame(lm_F)
res_final <- residuals(lm_F)

# Split main effects and interactions
main_terms <- all_terms[!grepl(":", all_terms)]
inter_terms <- all_terms[grepl(":", all_terms)]

# Build values for all terms
term_vals <- lapply(all_terms, function(term) {
  if (grepl(":", term)) {
    parts <- strsplit(term, ":")[[1]]
    mf_final[[parts[1]]] * mf_final[[parts[2]]]
  } else {
    mf_final[[term]]
  }
})
names(term_vals) <- all_terms

par(mfrow = c(ceiling(12 / 4), 4), mar = c(3, 3, 2, 1))
for (v in all_terms) {
  plot(term_vals[[v]], res_final,
       xlab = v, ylab = "Residuals",
       main = paste("Residuals vs", v),
       pch = 16, cex = 0.5, cex.main = 0.6,
       col = adjustcolor("steelblue", alpha.f = 0.5))
  abline(h = 0, col = "red", lty = 2)
}
par(mfrow = c(1, 1))
```

# 5. Extra Sum of Squares Test

The Extra Sum of Squares F-test formally answers our research question: **do the seven economic distress variables jointly provide statistically significant explanatory power for county-level Democratic vote share, beyond what traditional predictors already explain?**

**Hypotheses:**

$$H_0: \beta_{\text{Unemp}} = \beta_{\text{UnemplComp}} = \beta_{\text{EIC}} = 
\beta_{\text{ChildTax}} = \beta_{\text{Senior}} = \beta_{\text{SocSec}} = 
\beta_{\text{Pensions}} = \beta_{\text{Senior} \times \text{SocSec}} = 
\beta_{\text{EIC} \times \text{Black}} = \beta_{\text{EIC} \times \text{Hispanic}} = 
\beta_{\text{ChildTax} \times \text{Hispanic}} = 
\beta_{\text{Senior} \times \text{Bach}} = 
\beta_{\text{UnemplComp} \times \text{Migr}} = 0$$

$$H_A: \text{At least one of the above } \beta \neq 0$$

That is, under $H_0$, the economic distress block — including its seven main effects and six theoretically motivated interactions with demographic and prosperity variables — adds no explanatory power beyond Model R. Under $H_A$, at least one distress-related coefficient is nonzero, indicating that economic distress carries unique predictive signal for county-level Democratic vote share beyond traditional controls.

The ESS F-test therefore covers **13 additional parameters** (7 main effects + 
6 interactions), with numerator $df = 13$.

```{r ess-test, echo=FALSE}
ess_result <- anova(lm_R, lm_F)
print(ess_result)

# Extract key values for inline reporting
sse_R  <- ess_result$RSS[1]
sse_F  <- ess_result$RSS[2]
df_num <- ess_result$Df[2]
df_den <- ess_result$Res.Df[2]
f_stat <- ess_result$F[2]
p_val  <- ess_result$`Pr(>F)`[2]

cat("\n--- ESS F-Test Summary ---\n")
cat("SSE(Model R):", round(sse_R, 4), "\n")
cat("SSE(Model F):", round(sse_F, 4), "\n")
cat("Extra SS (SSE_R - SSE_F):", round(sse_R - sse_F, 4), "\n")
cat("Numerator df (number of distress predictors):", df_num, "\n")
cat("Denominator df (Model F residual df):", df_den, "\n")
cat("F-statistic:", round(f_stat, 4), "\n")
cat("p-value:", format(p_val, scientific = TRUE), "\n")
cat("final N =", nrow(df6))
```

**Interpretation:**



# 6. Conclusion


