---
title: "An Extra Sum of Stress: Does Economic Distress Predict Democratic Support Beyond General Prosperity and Demographic Composition?"
author: "Khanh Do, Maral Bat-Erdene"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggrepel)
library(readr)
library(tidyverse)
library(corrplot)
require(leaps)
library(patchwork)
require(mosaic)
require(corrplot)
require(leaps)
library(ggplot2)
library(reshape2)
```

# Read in Data

```{r message = FALSE, warning = FALSE, echo = FALSE}
url <- "https://raw.githubusercontent.com/khanhdo05/sta-310-final/main/data/clean/FINAL_DATA.csv"
df <- read_csv(url, show_col_types = FALSE)
```

Read csv from this [link](https://raw.githubusercontent.com/khanhdo05/sta-310-final/main/data/clean/FINAL_DATA.csv) into a dataframe `df`.

Code book available [here](https://docs.google.com/spreadsheets/d/1A-s4d9eTGT6rGAFhPsAnOutXermwbm6LxZr9JXc_NPE/edit?gid=0#gid=0).

Each row is unique by column `FIPS`.

# 1. Data Transformation

## Response Variable: PresDemRatio

```{r echo=FALSE}
hist(df$PresDemRatio,
     main = "Distribution of PresDemRatio",
     col = "steelblue")
```

The data looks right-skewed, but we shouldn't use log-transformation because this variable is a proportion bounded by 0 and 1. Log-transformation is too agressive for this data and can turn a right-skewed distribution to left-skewed. A square-root transformation is milder and more appropriate. We can see this more clearly with the histograms below.

```{r echo=FALSE}
hist(log(df$PresDemRatio),
     main = "Distribution of Log-Transformed PresDemRatio",
     col = "steelblue")
hist(sqrt(df$PresDemRatio),
     main = "Distribution of Sqrt-Transformed PresDemRatio",
     col = "steelblue")
```

The distribution of the square-root transformed one is not perfect but is definitely more bell-shape looking than that with log. Thus, we determine that our response variable needs a square-root transformation.

## Exploratory Variable

```{r}
# ---------------- PREP DATA ----------------
df_out <- df %>%
  select(-FIPS, -CountyName) %>%
  drop_na()

response <- "PresDemRatio"

predictors <- setdiff(names(df_out), response)

# ---------------- LOOP THROUGH VARIABLES ----------------
par(mfrow = c(1,1))  # reset plotting

for (var in predictors) {
  
  temp <- df_out %>%
    select(all_of(c(response, var))) %>%
    rename(x = all_of(var), y = all_of(response))
  
  # IQR outlier detection
  Q1 <- quantile(temp$x, 0.25, na.rm = TRUE)
  Q3 <- quantile(temp$x, 0.75, na.rm = TRUE)
  IQR_val <- Q3 - Q1
  
  temp <- temp %>%
    mutate(
      is_outlier = x < (Q1 - 1.5 * IQR_val) | x > (Q3 + 1.5 * IQR_val)
    )
  
  # ---------------- PRINT VARIABLE HEADER ----------------
  cat("\n---------------------------------------------\n")
  cat("Variable:", var, "\n")
  cat("Number of outliers:", sum(temp$is_outlier), "\n")
  
  # ---------------- PLOT ----------------
  p <- ggplot(temp, aes(x = x, y = y)) +
    geom_point(alpha = 0.5, color = "gray40") +
    
    # Highlight outliers
    geom_point(
      data = temp %>% filter(is_outlier),
      color = "red",
      size = 2
    ) +
    
    # Label extreme ones (optional but helpful)
    geom_text_repel(
      data = temp %>% filter(is_outlier) %>% head(5),
      aes(label = round(x, 2)),
      size = 3
    ) +
    
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    
    labs(
      title = paste("Outlier Detection:", var, "vs PresDemRatio"),
      x = var,
      y = "PresDemRatio"
    ) +
    theme_minimal()
  
  print(p)
}
```


```{r}

library(tidyverse)
library(corrplot)
library(leaps)
library(car)

# ---------------- PREP DATA ----------------
df <- df %>%
  select(-FIPS, -CountyName) %>%
  drop_na()

cat("Sample size:", nrow(df), "\n")

# ---------------- CORRELATION PLOT ----------------
corrplot(cor(df), 
         title = "Correlation Matrix (Final Data)", 
         mar = c(0,0,1,0), tl.cex = 0.5)

# ---------------- PAIRS PLOTS (GROUPED) ----------------

# Group 1: Income / Economic
pairs(df %>% select(PresDemRatio, TotalAgi, SalariesWages, BusinessNetIncome,
                    CapitalGains, UnemploymentComp, SocialSecurityBenefits),
      main = "Income Variables vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("steelblue", alpha.f = 0.4))

# Group 2: Education & Demographics
pairs(df %>% select(PresDemRatio, PctNoHs, PctHsOnly, PctSomeCollege,
                    PctBachelors, PctWhite, PctBlack, PctHispanic),
      main = "Education & Race vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("darkgreen", alpha.f = 0.4))

# Group 3: Population & Age
pairs(df %>% select(PresDemRatio, CensusPop2020, MedianAge, Pct18_34,
                    Pct65Plus, AgeDependency),
      main = "Population & Age vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("purple", alpha.f = 0.4))

# Group 4: Taxes & Wealth
pairs(df %>% select(PresDemRatio, TotalTaxLiability, MortgageInterest,
                    CharitableContributions, EicAmount, NumEicReturns),
      main = "Tax & Wealth vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("firebrick", alpha.f = 0.4))

# ---------------- MISSING DATA ----------------
missing_summary <- df %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing") %>%
  mutate(
    Total = nrow(df),
    Pct_Missing = round(100 * Missing / Total, 1)
  ) %>%
  arrange(desc(Missing))

print(missing_summary)

# ---------------- BEST SUBSETS ----------------
df$sq_PresDemRatio <- sqrt(df$PresDemRatio)

regfit <- regsubsets(sq_PresDemRatio ~ ., data = df, nvmax = 25)

plot(regfit, scale = "adjr2",
     main = "Best Subsets (Voting Model)",
     cex.axis = 0.6)

get_best <- function(model) {
  s <- summary(model)
  data.frame(
    NumVars = 1:length(s$adjr2),
    AdjR2 = s$adjr2,
    s$which
  ) %>% arrange(desc(AdjR2))
}

best_models <- get_best(regfit)
head(best_models, 10)

# ---------------- FIT MODEL ----------------
lm_best <- lm(sq_PresDemRatio ~ 
                TotalAgi +
                CapitalGains +
                UnemploymentRate2020 +
                PctBachelors +
                PctWhite +
                MedianAge +
                CensusPop2020 +
                TotalTaxLiability +
                EicAmount,
              data = df)

summary(lm_best)

# ---------------- RESIDUAL DIAGNOSTICS ----------------
par(mfrow = c(1,1))
plot(lm_best)

# ---------------- RESIDUALS VS EACH VARIABLE ----------------
vars <- names(df)
vars <- vars[vars != "sq_PresDemRatio"]

par(mfrow = c(1,1))

for (v in vars) {
  plot(df[[v]], lm_best$residuals,
       xlab = v,
       ylab = "Residuals",
       main = paste("Residuals vs", v),
       pch = 16, cex = 0.5)
  abline(h = 0, col = "red", lty = 2)
}
```

