---
title: "An Extra Sum of Stress: Does Economic Distress Predict Democratic Support Beyond General Prosperity and Demographic Composition?"
author: "Khanh Do, Maral Bat-Erdene"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggrepel)
library(readr)
library(tidyverse)
library(corrplot)
require(leaps)
library(patchwork)
require(mosaic)
require(corrplot)
require(leaps)
library(ggplot2)
library(reshape2)
library(car)
library(broom)
```

# Overview

## About our Data

```{r message = FALSE, warning = FALSE, echo = FALSE}
url <- "https://raw.githubusercontent.com/khanhdo05/sta-310-final/main/data/clean/FINAL_DATA.csv"
df <- read_csv(url, show_col_types = FALSE)
```

Read csv from this [link](https://raw.githubusercontent.com/khanhdo05/sta-310-final/main/data/clean/FINAL_DATA.csv) into a dataframe `df`.

Code book available [here](https://docs.google.com/spreadsheets/d/1A-s4d9eTGT6rGAFhPsAnOutXermwbm6LxZr9JXc_NPE/edit?gid=0#gid=0).

Each row is unique by column `FIPS`.

```{r echo=FALSE}
cat("Original data's nubmer of rows:", nrow(df), "\n")
cat("Original data's nubmer of columns:", ncol(df), "\n")
```

## Abstract

In this analysis, we are interested in exploring if the **Economic distress markers:**

- Unemployment Rate (UnemploymentRate2020)

- Reliance on unemployment compensation (PropUnemplComp)

- Breadth and intensity of low-income household prevalence (PropEicReturns and PropEicAmount)

- Elderly and aging population burdens (OldAgeDependency, PropSocSec, and PropPensions)

to provide unique explanatory power beyond traditional demographic, education, and prosperity indicators to significantly improve the prediction of the **proportion of Democrat votes in 2020 presidential election (PresDemRatio)**.

This investigation, rooted in the use of an Extra Sum of Squares test, seeks to determine if these specific indicators of community "stress" reduce model error and offer insights that general income levels alone cannot capture. 

Following the background research of Autor et al. (2020) and Gelman (2008), we aim to see if these localized economic conditions drive a measurable shift in county-level political identity.

# 1. Data Transformation

## 1.1 Response Variable: PresDemRatio

```{r echo=FALSE}
hist(df$PresDemRatio,
     main = "Distribution of PresDemRatio",
     col = "steelblue")
```

The data looks right-skewed, but we shouldn't use log-transformation because this variable is a proportion bounded by 0 and 1. Log-transformation is too aggressive for this data and can turn a right-skewed distribution to left-skewed. A square-root transformation is milder and more appropriate. We can see this more clearly with the histograms below.

```{r echo=FALSE}
hist(log(df$PresDemRatio),
     main = "Distribution of Log-Transformed PresDemRatio",
     col = "steelblue")
hist(sqrt(df$PresDemRatio),
     main = "Distribution of Sqrt-Transformed PresDemRatio",
     col = "steelblue")
```

The distribution of the square-root transformed one is not perfect but is definitely more bell-shape looking than that with log. Thus, we determine that our response variable needs a square-root transformation.

```{r echo=FALSE}
# Apply sqrt transformation to response
df1 <- df %>%
  mutate(sq_PresDemRatio = sqrt(PresDemRatio)) %>%
  relocate(sq_PresDemRatio) %>%
  select(-PresDemRatio)
```


## 1.2 Exploratory Variable

### 1.2.1 Remove Missing Data and Uninterested Columns

We first check for NAs to remove columns with a lot of NAs.

```{r echo=FALSE}
# ---------------- MISSING DATA ----------------
missing_summary <- df1 %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing") %>%
  mutate(
    Total = nrow(df),
    Pct_Missing = round(100 * Missing / Total, 1)
  ) %>%
  arrange(desc(Missing))

print(missing_summary)
```

None of our variables have more than 0.5% of missing data so we do not have to exclude any variables (remove no columns). However, the data includes both raw counts and some proportions and we would be more interested in the proportions than raw counts to reduce multicollinearity later on. We want to keep TotalAgi but that is the only raw count we are keeping.

```{r echo=FALSE}
df2 <- df1 %>%
  select(
    # remove raw dollar counts because they have proportion or average counterparts
    -SalariesWages, 
    -BusinessNetIncome, 
    -CapitalGains, 
    -PensionsAnnuities, 
    -UnemploymentComp, 
    -SocialSecurityBenefits, 
    -PartnershipIncome, 
    -MortgageInterest, 
    -CharitableContributions, 
    -TotalTaxLiability, 
    -EicAmount,
    -ChildTaxCredit,
    
    # remove raw head counts because they have proportion counterparts
    -NumEicReturns,
    -NetMigration2020,
    -CensusPop2020,
    -NumTaxReturns
  )
cat("Number of columns before removing raw counts:", ncol(df1), "\n")
cat("Number of rows after removing raw counts:", ncol(df2), "\n")
```

Then we remove rows with NAs.

```{r echo=FALSE}
# remove rows with NAs
df3 <- df2 %>%
  drop_na()
cat("Number of rows before removing rows with NAs:", nrow(df2), "\n")
cat("Number of rows after removing rows with NAs:", nrow(df3), "\n")
```


### 1.2.2 Outliers

```{r echo=FALSE}
# Prep data
df_out <- df3 %>%
  select(-FIPS, -CountyName) %>%
  drop_na()

response   <- "sq_PresDemRatio"
predictors <- setdiff(names(df_out), response)

# Create a list to store plots
plot_list <- list()

# Outlier detection loop
for (var in predictors) {
  
  temp <- df_out %>%
    select(all_of(c(response, var))) %>%
    rename(x = all_of(var), y = all_of(response))
  
  n <- nrow(temp)
  model   <- lm(y ~ x, data = temp)
  cooks_d <- cooks.distance(model)
  cutoff  <- 4 / n
  
  temp <- temp %>%
    mutate(cooks_d = cooks_d, is_outlier = cooks_d > cutoff)
  
  # Create the ggplot and save it to the list
  p <- ggplot(temp, aes(x = x, y = y)) +
    geom_point(aes(color = is_outlier), alpha = 0.4) +
    scale_color_manual(values = c("gray60", "red")) +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    labs(
      title = paste(var),
      subtitle = paste("Outliers:", sum(temp$is_outlier)),
      x = NULL, y = NULL
    ) +
    theme_minimal(base_size = 8) +
    theme(legend.position = "none")
  
  plot_list[[var]] <- p
}

# Display plots
wrap_plots(plot_list[1:9], ncol = 3, nrow = 3) +
  plot_annotation(title = "Outlier Detection 1")
wrap_plots(plot_list[10:18], ncol = 3, nrow = 3) +
  plot_annotation(title = "Outlier Detection 2")
wrap_plots(plot_list[19:27], ncol = 3, nrow = 3) +
  plot_annotation(title = "Outlier Detection 3")
wrap_plots(plot_list[28:36], ncol = 3, nrow = 3) +
  plot_annotation(title = "Outlier Detection 4")
```

Cook's Distance (Cook, 1977) is used here as the primary outlier detection method because it measures the influence of each observation on the fitted regression coefficients which is directly relevant to our goal of building a regression model. Unlike distributional methods such as IQR or Z-score outlier detection, which only flagged observations based on a single variable's distribution, Cook's D identifies observations whose removal would substantially change the estimated relationship between a predictor and the response variable (Belsley, Kuh & Welsch, 1980). The standard threshold of 4/n (Cook & Weisberg, 1982) is applied, where n is the number of complete observations for each predictor-response pair. Although the outliers are identified, we don't remove them without good reasons.

### 1.2.3 Correlation Plots

```{r echo=FALSE}
# Remove categorical/binary variables before computing correlation
df_corr <- df_out %>%
  select(-IsMetro) %>%
  select(where(is.numeric))

corrplot(cor(df_corr),
         title   = "Correlation Matrix",
         mar     = c(0, 0, 1, 0),
         tl.cex  = 0.5)
```
**Strong positive correlation:** The Democratic vote share is strongly positively correlated with education (PctBachelors), diversity (PctBlack, PctAsian, PctHispanic), and urbanization (PctUrban). Higher gross income (TotalAgi), proportion of income on mortgage interest (PropMortgage), average wages (AvgWage), tax liability (PropTaxLiability), unemployment compensation (PropUnemplComp), and charitable giving (PropCharity) are also positively correlated with sq_PresDemRatio. 

**Strong negative correlation:** There is a clear negative correlation with the white population percentage (PctWhite), high school only education (PctHsOnly), child dependency (ChildDependency), old age dependency (OldAgeDependency) and social security reliance (PropSocSec).

**Correlation with Voting:** Interestingly, several "stress" markers like PropEicReturns (breadth of poverty) show a negative correlation with sq_PresDemRatio. This aligns with the background research by Autor et al. (2020) cited in our proposal, which found that economic decline sometimes correlates with a shift toward Republican candidates rather than Democratic ones.

**Multicollinearity:** The matrix highlights several clusters where variables are so highly correlated that they are likely redundant:

- **Economic distress:** There are very large blue circles between PropEicAmount, PropEicReturns, and PropChildTax. This confirms that these variables are measuring similar aspects of community distress. Because they are so highly correlated, including all of them in a single model will likely trigger the high VIFs which will be addressed formally in part **1.2.4**. 

- **Education:** PctBachelors is strongly negatively correlated with PctNoHs and PctHsOnly. We will likely pick only one (PctBachelors) to represent education levels in our Reduced Model to avoid multicollinearity.

- **Age:** There is extreme correlation between MedianAge, Pct65Plus, and OldAgeDependency. We will use Best Subsets Regression later to determine which single age-related variable is the most efficient predictor.

- **Race:** PctWhite has prominent red circles at its intersections with PctBlack, PctHispanic, and PctAsian. This is expected in a closed system where these variables represent parts of a whole (100% of the population). We will Likely select PctWhite (as a baseline) and PctBlack and PctAsian (as markers of diversity)—rather than using the entire set.

- **Prosperity:** TotalAgi shows a positive relationship with AvgWage and PropCharity, but its correlation with the vote share is less pronounced than the "per-capita" or "proportion" versions. PropBusiness and AvgBusiness have strong correlation as well since they share the same numerator - we will likely just use PropBusiness because it aligns with our research into general prosperity.

```{r echo=FALSE}
# Group 1: General Prosperity and Affluence
pairs(df3 %>% select(sq_PresDemRatio, TotalAgi, PropCapitalGains, PropPartnership, 
                         PropMortgage, PropCharity, PropTaxLiability, PropSalaries, AvgWage),
      main = "General Prosperity vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("steelblue", alpha.f = 0.4))
```

**Group 1: General Prosperity and Affluence:** This group shows a strong "scale problem" where raw totals like TotalAgi are heavily skewed and highly correlated with each other. However, when normalized, markers like AvgWage and PropCharity show a positive correlation with the Democratic vote share, illustrating the "prosperity paradox" where wealthier counties often lean left. **Transformation Needed**: Log transformation for TotalAgi to handle extreme right-skewness and square root for prosperity ratios (like PropSalaries) to stabilize variance.

```{r echo=FALSE}
# Group 2: Economic Distress and Financial Vulnerability
pairs(df3 %>% select(sq_PresDemRatio, UnemploymentRate2020, PropUnemplComp, 
                         PropSocSec, PropPensions, PropEicAmount, PropEicReturns, PropChildTax),
      main = "Economic Distress vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("firebrick", alpha.f = 0.4))
```

**Group 2: Economic Distress and Financial Vulnerability:** There is significant internal redundancy among EIC-related variables and child tax credits. Interestingly, many markers of "stress" like PropEicReturns show a negative correlation with sq_PresDemRatio in these plots, suggesting that economic decline in certain areas may correlate with a move away from Democratic support. **Transformation Needed:** Square root for these ratios to handle mild skewness and boundary effects.

```{r echo=FALSE}
# Group 3: Educational Attainment
pairs(df3 %>% select(sq_PresDemRatio, PctNoHs, PctHsOnly, PctSomeCollege, PctBachelors),
      main = "Education Levels vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("darkgreen", alpha.f = 0.4))
```

**Group 3: Educational Attainment:** PctBachelors is one of the strongest positive predictors for Democratic support. The matrix reveals high multicollinearity within this group, particularly a strong negative relationship between PctBachelors and lower education levels like PctHsOnly. **Transformation Needed:** Square root for these percentages to ensure they meet the linearity assumptions of the regression model.

```{r echo=FALSE}
# Group 4: Racial and Gender Demographics
pairs(df3 %>% select(sq_PresDemRatio, PctWhite, PctBlack, PctAsian, PctHispanic, PctMale),
      main = "Demographics vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("darkorchid", alpha.f = 0.4))
```

**Group 4: Racial and Gender Demographics:** This group exhibits a "zero-sum" multicollinearity effect, where PctWhite is strongly negatively correlated with all other racial categories. Diversity markers (PctBlack, PctHispanic, PctAsian) are generally positive predictors for sq_PresDemRatio. **Transformation Needed:** Square root for all racial percentage variables.

```{r echo=FALSE}
# Group 5: Age and Dependency Ratios
pairs(df3 %>% select(sq_PresDemRatio, MedianAge, Pct18_34, Pct65Plus, 
                         AgeDependency, OldAgeDependency, ChildDependency),
      main = "Age and Dependency vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("darkorange", alpha.f = 0.4))
```

**Group 5: Age and Dependency Ratios:** There is a massive "super-cluster" of multicollinearity between MedianAge, Pct65Plus, and OldAgeDependency. These variables are virtually redundant, and the plots show that older counties generally correlate with lower Democratic vote shares. **Transformation Needed:** Square root for dependency ratios to stabilize the variance of these calculated proportions.

```{r echo=FALSE}
# Group 6: Urbanization and Population Dynamics
pairs(df3 %>% select(sq_PresDemRatio, PctUrban, IsMetro, BirthRate2020, 
                         DeathRate2020, MigrationRate2020),
      main = "Urbanization and Dynamics vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("dodgerblue", alpha.f = 0.4))
```

**Group 6: Urbanization and Population Dynamics:** PctUrban shows a distinct, non-linear positive relationship with Democratic voting. This group captures the urban-rural divide, with metropolitan integration (IsMetro) acting as a clear categorical separator. **Transformation Needed:** Square root for PctUrban to address its right-skewed distribution.

```{r echo=FALSE}
# Group 7: Business and Self-Employment Income
pairs(df3 %>% select(sq_PresDemRatio, PropBusiness, AvgBusiness),
      main = "Business Income vs Voting",
      pch = 16, cex = 0.3,
      col = adjustcolor("goldenrod", alpha.f = 0.4))
```

**Group 7: Business and Self-Employment Income:** PropBusiness and AvgBusiness are highly redundant. This group is unique because it can contain negative values (representing business losses), which prevents standard log transformations. **Transformation Needed:** An Inverse Hyperbolic Sine (IHS) transformation is required to handle negative values while mimicking a log scale to manage skewness (as established in our conversation history).

```{r echo=FALSE}
# ---------------- APPLY TRANSFORMATIONS ----------------
df4 <- df3 %>%
  mutate(
    # group 1
    log_TotalAgi = log1p(TotalAgi),
    sq_PropSalaries = sqrt(PropSalaries),
    sq_PropCharity = sqrt(PropCharity),
    sq_PropTaxLiability = sqrt(PropTaxLiability),
    
    # group 2
    sq_PropUnemplComp = sqrt(PropUnemplComp),
    sq_PropEicAmount = sqrt(PropEicAmount),
    sq_PropEicReturns = sqrt(PropEicReturns),
    sq_PropChildTax = sqrt(PropChildTax),
    
    # group 3
    sq_PctBachelors = sqrt(PctBachelors),
    sq_PctNoHs = sqrt(PctNoHs),
    sq_PctHsOnly = sqrt(PctHsOnly),
    
    # group 4
    sq_PctWhite = sqrt(PctWhite),
    sq_PctBlack = sqrt(PctBlack),
    sq_PctHispanic = sqrt(PctHispanic),
    sq_PctAsian = sqrt(PctAsian),
    
    # group 5
    sq_OldAgeDependency = sqrt(OldAgeDependency),
    sq_ChildDependency = sqrt(ChildDependency),
    sq_AgeDependency = sqrt(AgeDependency),
    
    # group 6
    sq_PctUrban = sqrt(PctUrban),
    sq_BirthRate = sqrt(BirthRate2020),
    sq_DeathRate = sqrt(DeathRate2020),
    
    # group 7
    ihs_PropBusiness = asinh(PropBusiness),
    ihs_AvgBusiness = asinh(AvgBusiness)
  ) %>%
  # Drop original untransformed versions only, keep everything else
  select(
    -TotalAgi, -PropSalaries, -PropCharity, -PropTaxLiability,
    -PropUnemplComp, -PropEicAmount, -PropEicReturns, -PropChildTax,
    -PctBachelors, -PctNoHs, -PctHsOnly,
    -PctWhite, -PctBlack, -PctHispanic, -PctAsian,
    -OldAgeDependency, -ChildDependency, -AgeDependency,
    -PctUrban, -BirthRate2020, -DeathRate2020,
    -PropBusiness, -AvgBusiness
  )

# ---------------- VERIFY TRANSFORMATIONS ----------------
par(mfrow = c(6, 4), mar = c(2, 2, 2, 1))
all_transformed_vars <- c(
  "log_TotalAgi", "sq_PropSalaries", "sq_PropCharity", "sq_PropTaxLiability",
  "sq_PropUnemplComp", "sq_PropEicAmount", "sq_PropEicReturns", "sq_PropChildTax",
  "sq_PctBachelors", "sq_PctNoHs", "sq_PctHsOnly",
  "sq_PctWhite", "sq_PctBlack", "sq_PctHispanic", "sq_PctAsian",
  "sq_OldAgeDependency", "sq_ChildDependency", "sq_AgeDependency",
  "sq_PctUrban", "sq_BirthRate", "sq_DeathRate",
  "ihs_PropBusiness", "ihs_AvgBusiness"
)

for (v in all_transformed_vars) {
  # Verify variable exists in df4 to avoid errors
  if (v %in% names(df4)) {
    hist(df4[[v]], 
         main = v, 
         xlab = "", 
         col = "steelblue", 
         border = "white",
         cex.main = 0.8) # Smaller title text to fit the grid
  } else {
    message(paste("Warning: Variable", v, "not found in df4."))
  }
}
par(mfrow = c(1, 1))

cat("Number of columns now:", ncol(df2), "\n")
cat("Number of rows now:", nrow(df2), "\n")
```

**Interpretation:**

log_TotalAgi is now roughly symmetric, correcting the extreme right-skew found in the raw income totals. Ratios such as sq_PropSalaries and sq_PropTaxLiability show significantly improved normality.

The "stress" markers (e.g., sq_PropUnemplComp, sq_PropEicAmount, sq_PropEicReturns) are no longer heavily clustered at the zero-boundary, which will allow your model to more accurately estimate the impact of financial vulnerability.

Variables like sq_PctBachelors and the age dependency ratios (sq_OldAgeDependency, sq_AgeDependency) now appear nearly normal. However, sq_PctBlack, sq_PctAsian, and sq_PctHispanic still show a right-skew because many U.S. counties have very small minority populations; the square root has "pulled" these values out to make them more manageable, but they remain inherently skewed.

sq_PctUrban still shows a spike at zero (representing entirely rural counties), but the overall distribution is more spread out than the raw percentage.

The IHS transformation (ihs_PropBusiness, ihs_AvgBusiness) successfully handled potential negative values while achieving a unimodal distribution.

### 1.2.4 Mean-centering and Multicollinearity (VIF)

We will first **mean-center** all numeric variables subtracting their mean. We exclude the unique identifier (FIPS), the categorical urbanization marker (IsMetro), and the transformed response variable (sq_PresDemRatio).

```{r echo=FALSE}
# ---------------- MEAN-CENTERING ----------------
# Mean-centering is essential before introducing interaction terms in Reduced Model
# to reduce multicollinearity between main effects and their interactions
# and to make intercepts interpretable

df5 <- df4 %>%
  mutate(across(where(is.numeric) & !any_of(c("FIPS", "sq_PresDemRatio", "IsMetro")), 
                ~ . - mean(., na.rm = TRUE)))

print("Verify mean is about 0:\n")
df5 %>%
  summarize(across(where(is.numeric), mean, na.rm = TRUE))
```

We then address the extreme redundancy seen in our Correlation Matrix through **VIFs**.

AgeDependency is the combination of OldAgeDependency and ChildDependency. Including all three is a "perfect" alias that makes VIF calculation mathematically impossible so we drop sq_AgeDependency. Like we mentioned earlier, we will use PropBusiness instead of AvgBusiness so we will drop ihs_AvgBusiness. We will also drop sq_PctNoHs. sq_PropEicAmount and sq_PropEicReturns is nearly identical and we will drop PropEicReturns since we are more interested in the intensity of poverty, not depth.

```{r echo=FALSE}
df6 <- df5 %>%
  select(-sq_AgeDependency, -ihs_AvgBusiness, -sq_PctNoHs, -sq_PropEicReturns)

cat("Number of colunms before dropping some variables:", ncol(df5), "\n")
cat("Number of colunms after dropping some variables:", ncol(df6), "\n")
```


```{r echo=FALSE}
# Run a preliminary model with ALL candidate predictors to identify high VIFs
vif_model <- lm(sq_PresDemRatio ~ . - FIPS - CountyName, data = df6)
vif_results <- vif(vif_model)
vif_df <- data.frame(Variable = names(vif_results), VIF = as.numeric(vif_results)) %>%
  arrange(desc(VIF))

# ---------------- FILTERED PRINTING ----------------
cat("Variables with VIF > 5 (Moderate Multicollinearity):\n")
print(vif_df %>% filter(VIF > 5 & VIF <= 10))

cat("Variables with VIF > 10 (SEVERELY Multicollinear - Must Prune):\n")
print(vif_df %>% filter(VIF > 10))
```

Our VIF analysis shows extreme multicollinearity in the full model, particularly within the age, dependency, and economic distress blocks. A VIF of 414 or 349 means these variables are almost perfectly redundant, which will inflate our standard errors and make our coefficient estimates unstable, potentially invalidating our Extra Sum of Squares (ESS) test. Thus, we must prune these variables to create a stable Reduced Model. The goal is to keep only the most theoretically important variable from each redundant cluster.

We will drop Pct65Plus (Redundant with sq_OldAgeDependency), MedianAge (Redundant with sq_OldAgeDependency), sq_ChildDependency (Redundant with sq_PropChildTax), sq_PctHsOnly (Redundant with sq_PctBachelors), sq_PctWhite (Redundant with diversity markers (zero-sum)), sq_PropChildTax.

```{r echo=FALSE}
# We remove the most redundant variables identified by the high VIFs
df_final <- df6 %>%
  select(
    -Pct65Plus,           # Redundant with sq_OldAgeDependency
    -MedianAge,           # Redundant with sq_OldAgeDependency
    -sq_ChildDependency,  # Redundant with sq_PropChildTax
    -sq_PctHsOnly,        # Redundant with sq_PctBachelors
    -sq_PctWhite,         # Redundant with diversity markers (zero-sum)
    -sq_PropChildTax      # Redundant with sq_PropEicAmount
  )

cat("Number of colunms before dropping some variables:", ncol(df6), "\n")
cat("Number of colunms after dropping some variables:", ncol(df_final), "\n")

# rerun VIF check
vif_model_v2 <- lm(sq_PresDemRatio ~ . - FIPS - CountyName, data = df_final)

vif_v2_results <- vif(vif_model_v2)

# Print results to ensure they are now mostly < 10
print("Check vifs again to ensure most are < 10")
vif_v2_df <- data.frame(Variable = names(vif_v2_results), VIF = as.numeric(vif_v2_results)) %>%
  arrange(desc(VIF))

print(vif_v2_df)
```


# 2. Best Subsets

We will now run Best Subsets Regression on our dataset (excluding the economic distress block) to select variables for our Reduced Model.

```{r echo=FALSE}
df_reduced_candidates <- df_final %>%
  select(
    -UnemploymentRate2020,
    -sq_PropUnemplComp,
    -sq_PropEicAmount,
    -sq_OldAgeDependency,
    -PropSocSec,
    -PropPensions
  )
```

```{r echo=FALSE}
fit_g1 <- regsubsets(
  sq_PresDemRatio ~ . -FIPS -CountyName,
  data   = df_reduced_candidates,
  nvmax  = 20,
  method = "seqrep"
)

s_g1 <- summary(fit_g1)

best_models_g1 <- data.frame(
  NumVars = 1:length(s_g1$adjr2),
  AdjR2   = round(s_g1$adjr2, 4),
  s_g1$which
) %>% arrange(desc(AdjR2))

# head(best_models_g1, 10)

best_n_g1 <- which.max(s_g1$adjr2)
cat("Best model size (Adj R2):", best_n_g1, "variables\n")
cat("Best Adj R2:", round(max(s_g1$adjr2), 4), "\n")

# Plot
plot(s_g1$adjr2, type = "b",
     xlab = "Number of Variables",
     ylab = "Adjusted R2",
     main = "Group 1 — Adjusted R2")
points(best_n_g1, max(s_g1$adjr2), col = "red", pch = 16, cex = 1.5)

plot(fit_g1, scale = "adjr2",
     main = "Variables Selection for Reduced Model",
     cex.axis = 0.55)

# Selected variables
vars_g1 <- names(coef(fit_g1, best_n_g1))[-1]
cat("Selected from the data:\n")
print(vars_g1)
```

## 3. Reduced Model

We will now make a reduced model **without interaction terms** first.

```{r echo=FALSE}
lm_reduced1 <- lm(sq_PresDemRatio ~ 
    PctMale + Pct18_34 + PropCapitalGains + PropPartnership + 
    PropMortgage + AvgWage + MigrationRate2020 + log_TotalAgi + 
    sq_PropSalaries + sq_PropCharity + sq_PropTaxLiability + sq_PctBachelors + 
    sq_PctBlack + sq_PctHispanic + sq_PctAsian + sq_PctUrban + 
    sq_DeathRate + ihs_PropBusiness + IsMetro, data=df_final)

cat("--- REDUCED MODEL WITHOUT INTERACTION SUMMARY --- \n")
summary(lm_reduced1)

par(mfrow = c(2, 2))
plot(lm_reduced1)
par(mfrow = c(1, 1))

# ---------------- RESIDUALS VS EACH PREDICTOR ----------------
model_df <- model.frame(lm_reduced1)
res <- residuals(lm_reduced1)
par(mfrow = c(ceiling(length(vars_plt) / 4), 4), mar = c(3, 3, 2, 1))
for (v in vars_g1) {
  plot(model_df[[v]], res,
       xlab = v,
       ylab = "Residuals",
       main = paste("Residuals vs", v),
       pch  = 16, cex = 0.5,
       col  = adjustcolor("steelblue", alpha.f = 0.5))
  abline(h = 0, col = "red", lty = 2)
}
par(mfrow = c(1, 1))
```

We will now test which interaction terms can be important.

```{r echo=FALSE}
lm_reduced2 <- lm(
  sq_PresDemRatio ~ 
    (PctMale + Pct18_34 + PropCapitalGains + PropPartnership + 
     PropMortgage + AvgWage + MigrationRate2020 + log_TotalAgi + 
     sq_PropSalaries + sq_PropCharity + sq_PropTaxLiability + 
     sq_PctBachelors + sq_PctBlack + sq_PctHispanic + 
     sq_PctAsian + sq_PctUrban + sq_DeathRate + 
     ihs_PropBusiness + IsMetro)^2,
  data = df_final
)

tidy_reduced <- tidy(lm_reduced2)

# ---------------- TOP 10 INTERACTIONS ----------------
top_interactions <- tidy_reduced %>%
  filter(grepl(":", term)) %>%      # interactions only
  filter(p.value < 0.001) %>%       # strong significance
  arrange(p.value) %>%              # smallest p-values first
  slice(1:10)                       # KEEP ONLY TOP 10

cat("\n--- TOP 10 INTERACTION TERMS ---\n")
print(top_interactions)

# ---------------- MAIN EFFECTS ----------------
significant_main <- tidy_reduced %>%
  filter(!grepl(":", term)) %>%
  filter(term != "(Intercept)") %>%
  filter(p.value < 0.05) %>%
  arrange(p.value)

cat("\n--- SIGNIFICANT MAIN EFFECTS ---\n")
print(significant_main)

# ---------------- BUILD FINAL TERM LIST ----------------
selected_terms <- c(
  significant_main$term,
  top_interactions$term
)

# ---------------- FINAL REDUCED MODEL ----------------
formula_reduced_final <- as.formula(
  paste(
    "sq_PresDemRatio ~",
    paste(selected_terms, collapse = " + ")
  )
)

lm_reduced_final <- lm(formula_reduced_final, data = df_final)

cat("\n--- FINAL REDUCED MODEL SUMMARY ---\n")
summary(lm_reduced_final)
par(mfrow = c(2, 2))
plot(lm_reduced_final)
par(mfrow = c(1, 1))

# ---------------- RESIDUALS VS EACH PREDICTOR ----------------
mf_final  <- model.frame(lm_reduced_final)
res_final <- residuals(lm_reduced_final)

# Split main effects and interactions
main_terms <- selected_terms[!grepl(":", selected_terms)]
inter_terms <- selected_terms[grepl(":", selected_terms)]

# Build values for all terms
term_vals <- lapply(selected_terms, function(term) {
  if (grepl(":", term)) {
    parts <- strsplit(term, ":")[[1]]
    mf_final[[parts[1]]] * mf_final[[parts[2]]]
  } else {
    mf_final[[term]]
  }
})
names(term_vals) <- selected_terms

par(mfrow = c(ceiling(length(vars_plt) / 4), 4), mar = c(3, 3, 2, 1))
for (v in selected_terms) {
  plot(term_vals[[v]], res_final,
       xlab = v, ylab = "Residuals",
       main = paste("Residuals vs", v),
       pch = 16, cex = 0.5, cex.main = 0.6,
       col = adjustcolor("steelblue", alpha.f = 0.5))
  abline(h = 0, col = "red", lty = 2)
}
par(mfrow = c(1, 1))
```

In the Residuals vs Fitted plot, observation 2636 is an extreme outlier. Let's investigate this.

```{r echo=FALSE}
df_final[2636,]
```

Loving County was identified as an influential observation with high leverage and large residual values in diagnostic plots. From online research, it is the least populous county in the United States, with a 2020 census population of 64. Due to its extremely small population and atypical predictor values, Loving County and Kalawao County of 82 population was removed to improve model stability and satisfy regression assumptions. Thus, we run our final Reduced Model again.

```{r echo=FALSE}
# Remove counties with CensusPop2020 < 100 from the df
small_counties <- df %>% filter(CensusPop2020 < 100) %>% select(FIPS, CountyName, CensusPop2020)
df_final_2 <- df_final %>% filter(!FIPS %in% small_counties$FIPS)
cat("Counties removed:\n")
print(small_counties)
cat("Counties remaining:", nrow(df_final_2), "\n")
```

```{r echo=FALSE}
lm_reduced3 <- lm(
  sq_PresDemRatio ~ 
    (PctMale + Pct18_34 + PropCapitalGains + PropPartnership + 
     PropMortgage + AvgWage + MigrationRate2020 + log_TotalAgi + 
     sq_PropSalaries + sq_PropCharity + sq_PropTaxLiability + 
     sq_PctBachelors + sq_PctBlack + sq_PctHispanic + 
     sq_PctAsian + sq_PctUrban + sq_DeathRate + 
     ihs_PropBusiness + IsMetro)^2,
  data = df_final_2
)

tidy_reduced <- tidy(lm_reduced3)

# ---------------- TOP 10 INTERACTIONS ----------------
top_interactions <- tidy_reduced %>%
  filter(grepl(":", term)) %>%      # interactions only
  filter(p.value < 0.001) %>%       # strong significance
  arrange(p.value) %>%              # smallest p-values first
  slice(1:10)                       # KEEP ONLY TOP 10

# ---------------- MAIN EFFECTS ----------------
significant_main <- tidy_reduced %>%
  filter(!grepl(":", term)) %>%
  filter(term != "(Intercept)") %>%
  filter(p.value < 0.05) %>%
  arrange(p.value)

# ---------------- BUILD FINAL TERM LIST ----------------
selected_terms <- c(
  significant_main$term,
  top_interactions$term
)

# ---------------- FINAL REDUCED MODEL ----------------
formula_reduced_final <- as.formula(
  paste(
    "sq_PresDemRatio ~",
    paste(selected_terms, collapse = " + ")
  )
)

lm_reduced_final <- lm(formula_reduced_final, data = df_final_2)

cat("\n--- FINAL REDUCED MODEL SUMMARY ---\n")
summary(lm_reduced_final)
par(mfrow = c(2, 2))
plot(lm_reduced_final)
par(mfrow = c(1, 1))

cat("Number of counties in the model:", nrow(df_final_2), "\n")

# ---------------- RESIDUALS VS EACH PREDICTOR ----------------
mf_final  <- model.frame(lm_reduced_final)
res_final <- residuals(lm_reduced_final)

term_vals <- lapply(selected_terms, function(term) {
  if (grepl(":", term)) {
    parts <- strsplit(term, ":")[[1]]
    mf_final[[parts[1]]] * mf_final[[parts[2]]]
  } else {
    mf_final[[term]]
  }
})
names(term_vals) <- selected_terms

par(mfrow = c(ceiling(length(vars_plt) / 4), 4), mar = c(3, 3, 2, 1))
for (v in selected_terms) {
  plot(term_vals[[v]], res_final,
       xlab = v, ylab = "Residuals",
       main = paste("Residuals vs", v),
       pch = 16, cex = 0.5, cex.main = 0.6,
       col = adjustcolor("steelblue", alpha.f = 0.5))
  abline(h = 0, col = "red", lty = 2)
}
par(mfrow = c(1, 1))
```

After removal of the influential outlier, diagnostic plots indicate that the 
assumptions of linearity, constant variance, normality, and absence of 
undue influence are reasonably satisfied. Minor deviations from normality 
remain in the upper tail, but these are not severe given the large sample size. Even though $R^2$ decreases, the removal is necessary to prevent our model from treating the extreme behaviors of 64 people as being as statistically significant as the trends seen in counties with millions of residents.

**Reduced Model Summary:**

Adjusted $R^2 = 0.667$
F-statistic p-value $< 2.2e-16$
Residual standard error $= 0.0776$

Thus, the model explains about 67% of variation in the response, which is strong for county-level social data.

## Full Model

(Need to get approval from Professor Kuiper first before building the Full Model and doing ESS)

Now that we have our the Reduced Model. We can build our Full Model.

## Extra Sum of Squares Test


## Conclusion


